---
title: "MSiA Bootcamp Final Exercise"
author: 'Zhonghao(Dennis) Zhao'
date: "Due Friday, Sept 18th 2020, 12:00 pm"
output:
  html_document:
    theme: readable
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# MSIA Boot Camp - Final R exercise

You've learned quite a lot about R in a short time. Congratulations! This exercise is designed to give you some additional practice on the material we have discussed this week while the lectures are still fresh in your mind, and to integrate different tools and skills that you have learned.

## Instructions

### Task 1: Import your data 

Read the data files `nys_schools.csv` and `nys_acs.csv` into R. These data come from two different sources: one is data on *schools* in New York state from the [New York State Department of Education](http://data.nysed.gov/downloads.php), and the other is data on *counties* from the American Communities Survey from the US Census Bureau. Review the codebook file so that you know what each variable name means in each dataset. 

```{r message = FALSE, warning = FALSE}
# load readr package
library(readr)
# import nys_schools data
nys_schools <- read_csv("/Users/zhonghaozhao/Desktop/MSiA_Bootcamp/R/bootcamp-2020/data/nys_schools.csv")
# import nys_acs data
nys_acs <- read_csv("/Users/zhonghaozhao/Desktop/MSiA_Bootcamp/R/bootcamp-2020/data/nys_acs.csv")
```

### Task 2: Explore your data

Getting to know your data is a critical part of data analysis. Take the time to explore the structure of the two dataframes you have imported. What types of variables are there? Is there any missing data? How can you tell? What else do you notice about the data?

```{r message = FALSE, warning = FALSE}
# check the number of null values in both dataset
sum(is.na(nys_schools))
sum(is.na(nys_acs))
# print summaries of 
str(nys_schools)
str(nys_acs)
```

The nyc_schools data have 1706 missing values. It also has 5 character value columns as well as 7 numerical value columns. The time period this data covers is from 2008-2017.


The nys_acs data have 0 missing values. It also has 1 character value as well as 4 numerical value columns. The time period this data covers is from 2009-2016.

### Task 3: Recoding and variable manipulation

1. Deal with missing values, which are currently coded as `-99`.
2. Create a categorical variable that groups counties into "high", "medium", and "low" poverty groups. Decide how you want to split up the groups and briefly explain your decision. 
3. The tests that the NYS Department of Education administers changes from time to time, so scale scores are not directly comparable year-to-year. Create a new variable that is the standardized z-score for math and English Language Arts (ELA) for each year (hint: group by year and use the `scale()` function)

```{r message = FALSE, warning = FALSE}
# substitue all -99 values to NA, then removes all rows containing missing values. 
nys_schools[nys_schools== -99]<-NA
nys_schools <- nys_schools[complete.cases(nys_schools),]

# get mean and sd of county_per_poverty to the county level
library(sqldf)
temp <- sqldf("select county_name, AVG(county_per_poverty) from nys_acs group by county_name")
m <- mean(temp$`AVG(county_per_poverty)`)
std <- sd(temp$`AVG(county_per_poverty)`)
# add the categorical column
temp$poverty_level <- cut(temp$`AVG(county_per_poverty)`, c(-Inf,m-std,m+std,Inf), c("low", "medium", "high"))
nys_acs <- sqldf("select * from nys_acs join temp on nys_acs.county_name = temp.county_name")
# remove unwanted columns after join
drops <- c("AVG(county_per_poverty)","county_name..6")
nys_acs <- nys_acs[ , !(names(nys_acs) %in% drops)]
#check column poverty_level added
names(nys_acs)

# group by year and take average ela and math scores
temp <- sqldf("select year, AVG(mean_ela_score) from nys_schools group by year")
annual_z_score_ela <- scale(temp)[,1]
nys_schools$annual_z_score_ela <- annual_z_score_ela[nys_schools$year-2007] 
temp <- sqldf("select year, AVG(mean_math_score) from nys_schools group by year")
annual_z_score_math <- scale(temp)[,1]
nys_schools$annual_z_score_math <- annual_z_score_math[nys_schools$year-2007]
#check new coluns annual_z_score_ela and annual_z_score_math added
names(nys_schools)

```

For task 3.1, I removed all rows containing -99 numerical values or NA values. 


For task 3.2, after aggregated the nys_acs data on county level, I classified the three group based on mean and standard deviation of the average county_per_poverty variable. Z score of -1 and 1 are the cut points. 


### Task 4: Merge datasets

Create a county-level dataset that merges variables from the schools dataset and the ACS dataset. Remember that you have learned multiple approaches on how to do this, and that you will have to decide how to summarize data when moving from the school to the county level.

```{r message = FALSE, warning = FALSE}
# check the number of null values in both dataset
# Join tables together
merged_df <- sqldf("SELECT *
               FROM nys_schools JOIN nys_acs 
               ON nys_schools.county_name = nys_acs.county_name and nys_schools.year = nys_acs.year")
# drop repeated columns
drops <- c("county_name..14","year..15")
merged_df <- merged_df[ , !(names(merged_df) %in% drops)]
names(merged_df)
```

I joined the two dataframes on the common columns year and county name.

### Task 5: Create summary tables

Generate tables showing the following:

1. For each county: total enrollment, percent of students qualifying for free or reduced price lunch, and percent of population in poverty.
2. For the counties with the top 5 and bottom 5 poverty rate: percent of population in poverty, percent of students qualifying for free or reduced price lunch, mean reading score, and mean math score.

```{r message = FALSE, warning = FALSE}
library(data.table)
# Task 5.1 Aggregate on county level from the merged dataframe
table_1 <- sqldf("SELECT county_name, SUM(total_enroll), AVG(per_free_lunch + per_reduced_lunch) as cheap_lunch,AVG(county_per_poverty) as poverty_rate
                 FROM merged_df
                 GROUP BY county_name")
# data table for the task
table_1 <- setDT(table_1) 

# Task 5.2 Aggregate on county level from the merged dataframe
temp <- sqldf("SELECT county_name, AVG(county_per_poverty) as poverty_rate,AVG(per_free_lunch + per_reduced_lunch) as cheap_lunch, AVG(mean_ela_score),AVG(mean_math_score)
                 FROM merged_df
                 GROUP BY county_name
                 ORDER BY poverty_rate")
# data table for the bottom 5 poverty rate
bottom_5 <- setDT(temp[1:5,]) 
bottom_5

top_5 <- sqldf("SELECT *
               FROM temp
               ORDER BY poverty_rate DESC")

# data table for the top 5 poverty rate
top_5 <- setDT(top_5[1:5,])
top_5
```

table_1, bottom_5, top_5 are three data tables answers the questions. The variable cheap_lunch is the averge of the sum of percentages of free lunch and reduced price lunch. It is used to show the percent of students qualifying for free or reduced price lunch.

### Task 6: Data visualization

Using `ggplot2`, visualize the following:

1. The relationship between access to free/reduced price lunch and test performance, at the *school* level.
2. Average test performance across *counties* with high, low, and medium poverty.

```{r message = FALSE, warning = FALSE}
# Task 6.1
library(ggplot2)
temp <- sqldf("SELECT school_name, AVG(per_free_lunch+per_reduced_lunch) as cheap_lunch, AVG(mean_ela_score) as read_scores, AVG(mean_math_score) as math_scores
              FROM merged_df
              GROUP BY school_name")

ggplot() + geom_point(aes(x=temp$cheap_lunch,y=temp$math_scores),color='red') +
geom_point(aes(x=temp$cheap_lunch,y=temp$read_scores),color='blue') +
ylab('Test score')+xlab('Percentage of cheap lunch')+ggtitle("Test Scores vs Lunch Prices")

# Task 6.2
read_score <- sqldf("SELECT poverty_level, AVG(mean_ela_score) as scores
              FROM merged_df
              GROUP BY poverty_level")
math_score <- sqldf("SELECT poverty_level, AVG(mean_math_score) as scores
              FROM merged_df
              GROUP BY poverty_level")

read_score$subject = rep("read_score",3)
math_score$subject = rep("math_score",3)
# modify df for plot purposes
temp <- rbind(read_score, math_score)
p <- ggplot(data=temp, aes(x=subject, y=scores, fill=poverty_level)) +
geom_bar(stat="identity", color="black", position=position_dodge())+
  theme_minimal()
# Use custom colors
p + scale_fill_brewer(palette="Blues") + ggtitle("Test Performance across Poverty Level")


```

### Task 7: Answering questions

Using the skills you have learned in the past three days, tackle the following question: 

> What can the data tell us about the relationship between poverty and test performance in New York public schools? Has this relationship changed over time? Is this relationship at all moderated by access to free/reduced price lunch?

You may use summary tables, statistical models, and/or data visualization in pursuing an answer to this question. Feel free to build on the tables and plots you generated above in Tasks 5 and 6.

```{r message = FALSE, warning = FALSE}
# check the number of null values in both dataset
# Join tables together
read_score <- sqldf("SELECT year, poverty_level, AVG(mean_ela_score) as scores
              FROM merged_df
              GROUP BY year, poverty_level")
math_score <- sqldf("SELECT year, poverty_level, AVG(mean_math_score) as scores
              FROM merged_df
              GROUP BY year, poverty_level")

# modify df for graph purposes
read_score$subject = rep("read_score",24)
math_score$subject = rep("math_score",24)

temp <- rbind(read_score, math_score)

# graphing with ggplot2
p <- ggplot(data=temp) +
geom_point(aes(x=year, y=scores, shape=subject, col=poverty_level)) +
ggtitle("Test Performance Over Time across Poverty Status")
p


```

Given the short time period, any answer will of course prove incomplete. The goal of this task is to give you some room to play around with the skills you've just learned. Don't hesitate to try something even if you don't feel comfortable with it yet. Do as much as you can in the time allotted.

**From the graph above in task 7, we can see that both math scores and read scores suffered a huge decrease in 2013. This is not normal. A potential reform may happened for both tests during that time. Besides the score drop across years, we can see a consistent pattern has exisited in all eight years. Regardless of the test subject, poverty level and test scores showed an inverse relationship with each other. Low poverty counties kept having higher scores than what the medium poverty counties had. Meanwhile, medium poverty counties were doing better in tests compared to high poverty counties. On the otherhand, according to our graph in task 6.1, there is no clear relationship between the access to free/reduced price lunch and test scores.**

## Github submission

When you have completed the exercise, save your Markdown file in the `submissions` folder of your forked repo using this naming convention: `FinalRExercise_LastnameFirstname.Rmd`. Commit changes periodically, and push commits when you are done.

You can optionally create a pull request to submit this file (and other exercise files from the bootcamp sessions) to the base repo that lives in the MSiA organization. If you would like to do this, make sure that all new files you have created are in the `submissions` folder, and then create a pull request that asks to merge changes from your forked repo to the base repo. 

## Reminders

- Remember to **load necessary packages**.
- Remember to **comment extensively** in your code. Since you will be working in an RMarkdown file, you can describe your workflow in the text section. But you should also comment within all of your code chunks.
- Attempt to knit your Markdown file into HTML format before committing it to Github. Troubleshoot any errors with the knit process by checking the lines referred to in the error messages.