---
title: "Final Exercise"
author: "Ben Hakimi"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Please make sure that you have downloaded both tidyverse and data.table
Both will be needed in this project and are loaded in here
```{r library}
library(tidyverse)
library(data.table)
```

Import New York Data Sets and save them for use in the project
```{r import data}
data_file1 <- here::here("data", "nys_schools.csv")
data_file2 <- here::here("data", "nys_acs.csv")

nys_schoolsdt <- fread(data_file1)

nys_acsdt <- fread(data_file2)

```


There are some missing values in the data which are currently coded in as 
"-99"
This will show us how many missing values are in each of our two data tables
``` {r null count}

sum(nys_schoolsdt == -99)

sum(nys_acsdt == -99)

```

Now that we know how we have missing values, we will remove the observations that have any missing values
```{r remove null}

nys_schoolsdt <- nys_schoolsdt[school_cd != -99 & 
                                school_name != -99 &
                                district_name != -99 &
                                county_name != -99 &
                                region != -99 &
                                year != -99 &
                                total_enroll != -99 &
                                per_free_lunch != -99 &
                                per_reduced_lunch != -99 &
                                per_lep != -99 &
                                mean_ela_score != -99 &
                                mean_math_score!= -99]

```

The variable "per_free_lunch" is a proportion and should only have values between 0 and 1.
Looking through the data however, I saw that there were some values greater than 1 for this variable 
so this command removes all observations with an "invalid" proportion for free lunch
```{r remove invalid}
nys_schoolsdt <- nys_schoolsdt[per_free_lunch <= 1]
```

This command splits the schools into 3 different poverty levels based on proportion
of students whi recieve a free lunch. by splitting the groups into

High: 55% - 100%
Med: 25% - 55%
Low: 0% - 25%

I found that it gives 3 fairly evenly sized groups that still sum up the data well
```{r new poverty variable}

nys_schoolsdt[per_free_lunch <= 1 & per_free_lunch >.55, pov_level := "high"]
nys_schoolsdt[per_free_lunch <= .55 & per_free_lunch >.25, pov_level := "med"]
nys_schoolsdt[per_free_lunch <= .25 & per_free_lunch >=0, pov_level := "low"]

```

Because the standardized tests vary year to year, this command breaks down the test scores into
standardized z-scores for each year of the data
``` {r create z score}
nys_schoolsdt[year == 2008, ela_z := (scale(nys_schoolsdt[year == 2008, mean_ela_score]))]
nys_schoolsdt[year == 2008, math_z := (scale(nys_schoolsdt[year == 2008, mean_math_score]))]

nys_schoolsdt[year == 2009, ela_z := (scale(nys_schoolsdt[year == 2009, mean_ela_score]))]
nys_schoolsdt[year == 2009, math_z := (scale(nys_schoolsdt[year == 2009, mean_math_score]))]

nys_schoolsdt[year == 2010, ela_z := (scale(nys_schoolsdt[year == 2010, mean_ela_score]))]
nys_schoolsdt[year == 2010, math_z := (scale(nys_schoolsdt[year == 2010, mean_math_score]))]

nys_schoolsdt[year == 2011, ela_z := (scale(nys_schoolsdt[year == 2011, mean_ela_score]))]
nys_schoolsdt[year == 2011, math_z := (scale(nys_schoolsdt[year == 2011, mean_math_score]))]

nys_schoolsdt[year == 2012, ela_z := (scale(nys_schoolsdt[year == 2012, mean_ela_score]))]
nys_schoolsdt[year == 2012, math_z := (scale(nys_schoolsdt[year == 2012, mean_math_score]))]

nys_schoolsdt[year == 2013, ela_z := (scale(nys_schoolsdt[year == 2013, mean_ela_score]))]
nys_schoolsdt[year == 2013, math_z := (scale(nys_schoolsdt[year == 2013, mean_math_score]))]

nys_schoolsdt[year == 2014, ela_z := (scale(nys_schoolsdt[year == 2014, mean_ela_score]))]
nys_schoolsdt[year == 2014, math_z := (scale(nys_schoolsdt[year == 2014, mean_math_score]))]

nys_schoolsdt[year == 2015, ela_z := (scale(nys_schoolsdt[year == 2015, mean_ela_score]))]
nys_schoolsdt[year == 2015, math_z := (scale(nys_schoolsdt[year == 2015, mean_math_score]))]

nys_schoolsdt[year == 2016, ela_z := (scale(nys_schoolsdt[year == 2016, mean_ela_score]))]
nys_schoolsdt[year == 2016, math_z := (scale(nys_schoolsdt[year == 2016, mean_math_score]))]

nys_schoolsdt[year == 2017, ela_z := (scale(nys_schoolsdt[year == 2017, mean_ela_score]))]
nys_schoolsdt[year == 2017, math_z := (scale(nys_schoolsdt[year == 2017, mean_math_score]))]
```

This merges the two data sets to get statistics per school and counties as a whole
```{r merge data}
county_data <- merge(nys_schoolsdt, nys_acsdt, by = c("county_name", "year"))
```

This command gives shows the total student enrollment, proportion of students recieveing
reduced/free lunch and the poverrty rate for each county over the time of the data
```{r county summary}

counties <- matrix(c(unique(county_data$county_name)))

for (i in counties){
countysummary <- county_data[county_name == i, .(enrollment = sum(total_enroll), 
                                      qualify = sum((per_free_lunch * total_enroll) +                                       (per_reduced_lunch * total_enroll)),                                                 percent_poverty = mean(county_per_poverty))]
  print(paste0("County: ", i))
  print(paste0("Total Enrollment   Qualifiers          Poverty Rate"))
  print(paste0(countysummary))
}    
```

This command singles out the poverty levels in 2016 for each county so we can use
it to break down the top and bottom five counties in terms of poverty levels
```{r sort poverty levels}
lowpovsort <- matrix(c(sort(nys_acsdt[year == 2016, county_per_poverty])))
highpovsort <- matrix(c(sort(nys_acsdt[year == 2016, county_per_poverty], decreasing = TRUE)))
```

This command singles out the top 5 and bottom 5 values of poverty rate in 2016
```{r get top and bottom 5 values}
lowsort5 <- matrix(c(lowpovsort[1:5,]))
highsort5 <- matrix(c(highpovsort[1:5,]))
```

This command pulls out the county names for the poverty levels obtained in the previous command.
This will allow us to pull out all necessary data for selected county
```{r get 5 names}
lowsortnames <- matrix(c(nys_acsdt[county_per_poverty %in% lowsort5, county_name]))
highsortnames <- matrix(c(nys_acsdt[county_per_poverty %in% highsort5, county_name]))
```

This command creates tables showing the Percent in Poverty, Proportion of county students 
that qualify for free/reduced price lunch, and average test scores across each county in 2016
for the top 5 and bottom five counties in terms of poverty percentage
```{r create top and bottom 5 tables}

low5data <- county_data[county_name %in% lowsortnames & year == 2016, .(Percent_in_Poverty = mean(county_per_poverty), Qualify = mean(per_free_lunch + per_reduced_lunch), Math_Score = mean(mean_math_score), ELA_Score = mean(mean_ela_score)), by = .(County = county_name)]

high5data <- county_data[county_name %in% highsortnames & year == 2016, .(Percent_in_Poverty = mean(county_per_poverty), Qualify = mean(per_free_lunch + per_reduced_lunch), Math_Score = mean(mean_math_score), ELA_Score = mean(mean_ela_score)), by = .(County = county_name)]

```

This command prints the tables we just created
```{r print tables}

low5data

high5data

```

This creates a scatterplot, with a point for each school, showing the relationship between 
reduced/free luch qualification rate and math test scores in 2017.

It shows that there is a clear negative relationship between the two meaning as 
qualification rate goes up, test scores apear to be lower.

Because higher qualification rate is associated with a higher poverty percentage this gives insight into 
test scores in lower income schools
```{r scores v lunch scatterplot}

nys_schoolsdt[year == 2017, .(qualify = per_free_lunch + per_reduced_lunch, score = mean_math_score)] %>% 
ggplot() + 
geom_point(mapping = aes(x=qualify, y = score)) +
labs(title = "Free/Reduced Lunch Qualification vs. Math Test Scores",
     subtitle = "In the year 2017",
     x = "Qualification Rate",
     y = "Test Score")

```

This again splits up poverty levels into 3 categories this time using the 
county percentage in poverty rate instead of school lunch data
```{r create level data}

county_data[county_per_poverty <= .3 & county_per_poverty >.16, pov_level2 := "high"]
county_data[county_per_poverty <= .16 & county_per_poverty >.12, pov_level2 := "med"]
county_data[county_per_poverty <= .12 & county_per_poverty >=0, pov_level2 := "low"]

```

This creates a boxplot for county test score split up into the high, medium, 
and low catergories we just created
```{r create boxplots}

unique(county_data[year == 2016, .(score = mean(mean_math_score), level = pov_level2), by = .(county = county_name)]) %>% 
ggplot() +
geom_boxplot(aes(x=level, y=score)) + 
labs(title = "Math Test Scores by Poverty Level",
     subtitle = "In the year 2016",
     x = "Poverty Level",
     y = "Test Score")

```

This creates a linear regression showing the rlationship between mean math test scores
and total enrolment in schools + county poverty level

This allows us to asses the impact on larger school size as well as poverty on test scores
```{r create regression}
mod = lm(mean_math_score ~ total_enroll + county_per_poverty, data = county_data)
summary(mod)
```

This creates a scaterplot, with a point for each county, showing the relationship between 
reduced/free luch qualification rate and math test scores in 2017

The data is the color coated to split the points up between high, medium, and low poverty levels
```{r county pov level scatterplot}

nys_schoolsdt[year == 2017, .(qualify = mean(per_free_lunch + per_reduced_lunch), score = mean(mean_math_score), Poverty_Level = pov_level), by = county_name] %>% 
ggplot() + 
geom_point(mapping = aes(x=qualify, y = score, col = Poverty_Level)) +
labs(title = "Free/Reduced Lunch Qualification vs. Math Test Scores",
     subtitle = "By County / In the year 2017",
     x = "Qualification Rate",
     y = "Test Score")

```


