---
title: "Exercises Day 8"
author: "Sara Ho"
date: "`r Sys.Date()`"
output: html_document
---

```{r global_options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE)
```

## 1. Import data and packages 

Import necessary packages.
```{r results = "hide"}
library(tidyverse)
library(data.table)
```

Load the data
```{r}
schools <- fread(here::here("data", "nys_schools.csv"))
counties <- fread(here::here("data", "nys_acs.csv"))
```

## 2. Explore and clean data

```{r}
summary(schools)
```

Deal with missing values, which are currently coded as `-99`.
```{r}
# since we need county level data later, let's remove schools without a county
schools <- schools[county_name != -99]

# for remaining columns, convert -99 to NA
metric_names <- c("total_enroll", "per_free_lunch", "per_reduced_lunch", "per_lep", "mean_ela_score", "mean_math_score")
schools[, (metric_names) := replace(.SD, .SD == -99, NA)
        , .SDcols = metric_names]

# verify here that there are no -99 in the minimum values
summary(schools)
```

Since these columns cannot be over 100%, convert numbers above 1 to fractions. If there are remaining columns still over 1, replace these data points with `NA`.
```{r}
schools[per_free_lunch > 1, per_free_lunch := per_free_lunch / 100]
schools[per_free_lunch > 1, per_free_lunch := NA]

schools[per_reduced_lunch > 1, per_reduced_lunch := per_reduced_lunch / 100]
schools[per_reduced_lunch > 1, per_reduced_lunch := NA]

# verify these variables are no larger than 1
summary(schools[, per_reduced_lunch])
summary(schools[, per_free_lunch])
```

### Recode `county` poverty variable

Create a categorical variable `pov_cat` that groups counties into "high", "medium", and "low" poverty groups.
For each year, designate the 25th percentile as the cutoff for low poverty, and the 75th percentile as the cutoff for high poverty. 

First, obtain the low and high cutoffs for each year.
```{r}
# `poverty_cutoffs` contains the low and high cutoffs for each year in `counties`
poverty_cutoffs <- counties[ , .(cutoff_low = quantile(county_per_poverty, 0.25), 
                            cutoff_high = quantile(county_per_poverty, 0.75)), year]
poverty_cutoffs
```

Then use the cutoffs to create `pov_cat`.
```{r}
# merge `poverty_cutoffs` with `counties`
counties <- counties[poverty_cutoffs, on = "year"]

# create the `pov_cat` variable
counties[, pov_cat := "medium"]
counties[county_per_poverty < cutoff_low, pov_cat := "low"]
counties[county_per_poverty > cutoff_high, pov_cat := "high"]

# delete the extra cutoff variables
counties[, cutoff_low := NULL][, cutoff_high := NULL]
head(counties)
```

### Create helpful additional variables in `school`

The tests that the NYS Department of Education administers changes from time to time, so scores are not directly comparable year-to-year. Create a new variable that is the standardized z-score for math and English Language Arts (ELA) for each year.

```{r}
schools[, z_mean_ela_score := scale(mean_ela_score), year]
schools[, z_mean_math_score := scale(mean_math_score), year]
```

Create variables `num_free_lunch` and `num_reduced_lunch` that represent the total number of students in free and reduced lunch.
```{r}
schools[ , num_free_lunch := round(total_enroll * per_free_lunch)]
schools[ , num_reduced_lunch := round(total_enroll * per_reduced_lunch)]
```

Also, create a variable `per_free_reduced_lunch` that sums the percentages of students in free lunch programs and students in reduced lunch programs
```{r}
# create `per_free_reduced_lunch`
schools[, per_free_reduced_lunch := per_free_lunch + per_reduced_lunch]

# For some schools, `per_free_reduced_lunch` is greater than 1.
head(schools[per_free_reduced_lunch > 1, .(per_free_reduced_lunch, per_free_lunch, per_reduced_lunch)])

# it's possible that students in the reduced lunch category here are included in the "free lunch" category.
# so let's replace `per_free_reduced_lunch` with `per_free_lunch` for these rows. 
schools[per_free_reduced_lunch > 1, per_free_reduced_lunch := per_free_lunch]
summary(schools$per_free_reduced_lunch)
```

### Merge datasets

Create a county-level data set that merges variables from the schools dataset and the ACS dataset. 

```{r}
merged <- merge(schools, counties, by = c("county_name", "year"), all.x = TRUE)
```

---

## 3. Create summary tables

**For each county: total enrollment, percent of students qualifying for free or reduced price lunch, and percent of population in poverty.**

First, get the number of total students per county enrolled in free or reduced lunch program:

```{r}
# mean county poverty across the years for which there is data
sum_table_county <- merged[ , .(tot_enroll = sum(total_enroll, na.rm = TRUE),
                           tot_reduced_lunch = sum(num_reduced_lunch, na.rm = TRUE),
                           tot_free_lunch = sum(num_free_lunch, na.rm = TRUE),
                           mean_poverty = mean(county_per_poverty, na.rm = TRUE)), county_name]
```

Then, convert the total numbers to percentages (fractions)

```{r}
sum_table_county[, p_reduced_lunch := tot_reduced_lunch / tot_enroll]
sum_table_county[, p_free_lunch := tot_free_lunch / tot_enroll]
sum_table_county[, tot_reduced_lunch := NULL][, tot_free_lunch := NULL]
head(sum_table_county)
```

**For the counties with the top 5 and bottom 5 poverty rate: percent of population in poverty, percent of students qualifying for free or reduced price lunch,mean reading score, and mean math score.**

We'll do this for the most current year with county data, which is 2016

First, select the counties with the top 5 and bottom 5 poverty rates from 2016, the most recent year.
```{r}
# slice the county dataset into `low_pov_counties` and `high_pov_counties`, each contain 5 counties
low_pov_counties <- counties[year == max(year)][order(county_per_poverty)] %>% slice(1:5)
high_pov_counties <- counties[year == max(year)][order(county_per_poverty, decreasing = TRUE)] %>% slice(1:5)
```

Scores are not comparable from year to year, but they should be comparable from school to school as long as the years are the same!

Check the data

**Low poverty schools**
```{r}
low_pov_schools <- schools[low_pov_counties, on = c("county_name", "year")]
summary(low_pov_schools$mean_ela_score)
summary(low_pov_schools$mean_math_score)
n_distinct(low_pov_schools$school_cd)
```

**High poverty schools**
```{r}
high_pov_schools <- schools[high_pov_counties, on = c("county_name", "year")]
summary(high_pov_schools$mean_ela_score)
summary(high_pov_schools$mean_math_score)
n_distinct(high_pov_schools$school_cd)
```

From this information, we know that:

* From the bottom 5 low poverty counties, there are 642 schools, of which 67 are missing ELA and math scores.
* From the top 5 high poverty counties, there are 831 schools, of which 38 are missing ELA and math scores.

It looks like missing score data is not correlated with high poverty levels.

Create a summary table for the schools from low poverty counties and a summary table for the schools from high poverty counties.
```{r}
# create low poverty summary table
sum_table_low_pov <- low_pov_schools[ , .(tot_enroll = sum(total_enroll, na.rm = TRUE),
                     tot_reduced_lunch = sum(num_reduced_lunch, na.rm = TRUE),
                     tot_free_lunch = sum(num_free_lunch, na.rm = TRUE),
                     mean_poverty = mean(county_per_poverty, na.rm = TRUE),
                     mean_math_score = mean(mean_math_score, na.rm = TRUE),
                     mean_ela_score = mean(mean_ela_score, na.rm = TRUE))
                 , county_name][, pov_cat := "low"]

# create high poverty summary table
sum_table_high_pov <- high_pov_schools[ , .(tot_enroll = sum(total_enroll, na.rm = TRUE),
                      tot_reduced_lunch = sum(num_reduced_lunch, na.rm = TRUE),
                      tot_free_lunch = sum(num_free_lunch, na.rm = TRUE),
                      mean_poverty = mean(county_per_poverty, na.rm = TRUE),
                      mean_math_score = mean(mean_math_score, na.rm = TRUE),
                      mean_ela_score = mean(mean_ela_score, na.rm = TRUE))
                  , county_name][, pov_cat := "high"]

# convert totals to percentages
sum_table_low_pov[, p_free_reduced_lunch := (tot_reduced_lunch + tot_free_lunch)/ tot_enroll]
sum_table_high_pov[, p_free_reduced_lunch := (tot_reduced_lunch + tot_free_lunch)/ tot_enroll]

# remove the columns with total student numbers
sum_table_low_pov[, tot_enroll := NULL][, tot_reduced_lunch := NULL][, tot_free_lunch := NULL]
sum_table_high_pov[, tot_enroll := NULL][, tot_reduced_lunch := NULL][, tot_free_lunch := NULL]
```

Here are the resulting summary tables:
```{r}
sum_table_low_pov
sum_table_high_pov
```


---

## 4.1 Data visualization and analysis (lunch programs)

> What can the data tell us about the relationship between poverty and test performance in New York public schools? Has this relationship changed over time? Is this relationship at all moderated by access to free/reduced price lunch?

```{r}
# set all ggplots to the same theme
theme_set(theme_light())
```

Create a general scatter plot function to re-use later

```{r}
# create a function called `scatter_plot` that takes data and a mapping as parameters
# the function only plots observations in the data with no missing data
# this should not be a problem since the future code only selects relevant columns
# the plot contains points, trend lines, and a simplified legend
scatter_plot <- function(plot_data, map){
  ggplot(plot_data[complete.cases(plot_data), ], mapping = map) + 
    geom_point(size = 0.3, alpha = 0.5) + 
    geom_smooth(method = "lm", se = FALSE) + 
    theme(legend.title = element_blank())
}
```

**Plot the relationship between access to free/reduced price lunch and test performance for 2016.** Each point corresponds to a school.

```{r}
# `model_data` contains a subset of the original schools data
# convert percentage to basis points for easy interpretation
model_data <- schools[year == max(year), .(school_name, school_cd, mean_ela_score, mean_math_score, per_free_reduced_lunch = per_free_reduced_lunch * 100)]

# reshape the data from wide to long for plotting purposes
plot_data <- melt(model_data, id.vars = c("school_name", "school_cd", "per_free_reduced_lunch"))

scatter_plot(plot_data, aes(x = per_free_reduced_lunch, y = value, color = variable)) + 
  scale_color_manual(values = c("blue", "darkred"), labels = c("ELA", "math")) +
  labs(title = "Percent of students receiving free or reduced lunch v test scores - 2016",
       x = "Students receiving free or reduced lunch (%)",
       y = "Scores")
```

Is free/reduced lunch a significant predictor of ELA scores?

```{r}
model = lm(formula = mean_ela_score ~ per_free_reduced_lunch, data = model_data)
summary(model)
```

Is free/reduced lunch a significant predictor of math scores?

```{r}
model = lm(formula = mean_math_score ~ per_free_reduced_lunch, data = model_data)
summary(model)
```

> In 2016, for every basis point increase in students enrolled in free or reduced lunch, ELA scores decrease by 0.35 points and math scores decrease by 0.45 points.

This does not mean that free and reduced lunch programs **cause** a decrease in scores; rather students who need the free and reduced lunch programs tend to have lower scores. Does a *change* in the use of programs result in a change of scores?

Let's compare the difference in use of lunch programs with difference in scores from year to year.

```{r}
# create a copy of `schools` so that we preserve the original data
# convert percentage to basis points for easy interpretation
model_data <- schools[, .(year, school_name, school_cd, z_mean_ela_score, z_mean_math_score, per_free_reduced_lunch = per_free_reduced_lunch * 100)]

# order `schools_plot` by school and year before creating difference variables
# because we are comparing scores year by year, we have to use the z-scores instead of the actual scores
setorder(model_data, school_cd, year)
model_data[, diff_lunch := per_free_reduced_lunch - shift(per_free_reduced_lunch), by = school_cd]
model_data[, diff_ELA_score := z_mean_ela_score - shift(z_mean_ela_score), by = school_cd]
model_data[, diff_math_score := z_mean_math_score - shift(z_mean_math_score), by = school_cd]

# reshape the data from wide to long for plotting multiple categories
plot_data <- melt(model_data[, .(school_name, school_cd, year, diff_lunch, diff_ELA_score, diff_math_score)], 
                  id.vars = c("school_name", "school_cd", "year", "diff_lunch"))

scatter_plot(plot_data, aes(x = diff_lunch, y = value, color = variable)) + 
  scale_color_manual(values = c("blue", "darkred"), labels = c("ELA", "math")) +
  labs(title = "Year-to-year change in test scores and lunch programs", 
       y = "Difference in score", 
       x = "Difference in percent receiving free or reduced lunch")
```

Is a change in free/reduced lunch a significant predictor a change in ELA scores?

```{r}
model = lm(formula = diff_ELA_score ~ diff_lunch, data = model_data)
summary(model)
```

Is a change in free/reduced lunch a significant predictor a change in math scores?
```{r}
model = lm(formula = diff_math_score ~ diff_lunch, data = model_data)
summary(model)
```

> For every basis point increase in students enrolled in free or reduced lunch from year-to-year, ELA scores decrease by 0.11 points and math scores decrease by 0.04 points. The effect on ELA scores is significant, but the effect on math scores is not.

There are two possible drivers for the change in program enrollment

1. An increased **access** of the program to students in need. If scenario is true, then our results show that an increase in access does not help to increase test scores
2. An increased number of students in need. If scenario is true, then our results only show us what we already know from the previous exercise: schools with a large percentage of students in free or reduced lunch programs tend to have lower test scores. Unfortunately this does not tell us whether the accessibility of the lunch program has an effect on test scores.

Unfortunately, without more granular data, we do not know which scenario is more accurate (could be both!).

To diagnose which scenario is more accurate, we would want to know whether the same students are represented in the data from year to year and whether their ability to afford lunch stayed constant from year to year.

We assume an increased access to affordable lunch has an effect on test scores over the same year, however, it may be true that an effect on test scores does not show up until *more* than a year of continued access to affordable lunch for students in need. 

---

## 4.2 Data visualization and analysis (poverty rate)

Average test performance across *counties* with high, low, and medium poverty.

```{r}
# `plot_data` contains a subset of the original merged data
plot_data <- merged[ , .(mean_ela = mean(z_mean_ela_score, na.rm = TRUE),
                         mean_math = mean(z_mean_math_score, na.rm = TRUE)), pov_cat]

# reshape the data from wide to long for plotting purposes
plot_data <- melt(plot_data, id.vars = "pov_cat")
# remove missing data
plot_data <- plot_data[complete.cases(plot_data), ]

ggplot(plot_data[!is.na(pov_cat)]) + 
  geom_col(aes(x = pov_cat, y = value, fill = variable)) + 
  scale_fill_manual(values = c("blue", "darkred")) +
  facet_grid( ~variable) + 
  labs(title = "Average test scores by poverty level", y = "Scores", x = "Poverty") + 
  theme(strip.background = element_blank(),
        panel.border = element_blank(),
        axis.ticks = element_blank(),
        panel.grid.major.x=element_blank(),
        legend.title = element_blank())
```

There is a clear difference between the scores from high poverty counties compared to the scores from low poverty counties.
Let's again use a scatter plot to visualize the relationship.

```{r}
# `model_data` contains a subset of the original merged data
# convert percentage to basis points for easy interpretation
model_data <- merged[year == 2016, .(school_name, school_cd, mean_ela_score, mean_math_score, county_per_poverty = county_per_poverty * 100)]

# reshape the data from wide to long for plotting purposes
plot_data <- melt(model_data, id.vars = c("school_name", "school_cd", "county_per_poverty"))

scatter_plot(plot_data, aes(x = county_per_poverty, y = value, color = variable)) + 
  scale_color_manual(values = c("blue", "darkred"), labels = c("ELA", "math")) +
  labs(title = "County-level poverty v school test scores - 2016",
       x = "County-level poverty (%)",
       y = "Scores")
```

The two groups on the far right with the highest levels of poverty are Bronx County and Kings County (Brooklyn).

```{r}
model = lm(formula = mean_ela_score ~ county_per_poverty, data = model_data)
summary(model)
```
There is a significant relationship between poverty rates and test scores. However, we see a very poor fit (low R2). It looks like there is a large variance of average scores across schools in each county. Confirm whether this is true in the data.

```{r}
# create a summary table with the standard deviations of scores across counties
# include the mean of scores and the total enrollment across counties for reference
score_sd <- merged[year == 2016, .(mean_ela = mean(mean_ela_score, na.rm = TRUE),
                                   mean_math = mean(mean_math_score, na.rm = TRUE),
                                   std_ela = sd(mean_ela_score, na.rm = TRUE),
                                   std_math = sd(mean_math_score, na.rm = TRUE), 
                                   total_enroll = sum(total_enroll, na.rm = TRUE),
                                   county_per_poverty = mean(county_per_poverty)), county_name]

# low poverty schools:
score_sd[order(county_per_poverty)] %>% slice(1:10)

# high poverty schools:
score_sd[order(county_per_poverty, decreasing = TRUE)] %>% slice(1:10)
```
In particular, the Bronx, which has the highest poverty rate, there is an average of 13 point variation from the mean across schools' ELA test scores, and an average of 18 point variation from the mean across schools' math test scores.

Across all schools, math scores have a higher variance than ela scores. There doesn't seem to be a difference in variance across poverty levels. Rather, variance has more to do with total enrollment. This makes sense - counties with a large number of students are likely to have more diversity in test results. Notably Kings County (Brooklyn) and New York County (Manhattan) both have very large variances.

```{r}
# reshape the data from wide to long for plotting purposes
plot_data <- melt(score_sd[,.(county_name, std_ela, std_math, total_enroll)], id.vars = c("county_name", "total_enroll"))

scatter_plot(plot_data, aes(x = total_enroll, y = value, color = variable)) + 
  scale_color_manual(values = c("blue", "darkred"), labels = c("ELA", "math")) +
  labs(title = "County-level total enrollment v test score standard deviation - 2016",
       x = "Total enrollment",
       y = "Test score sd")
```
```{r}
low_enroll_counties <- score_sd[total_enroll <= quantile(score_sd$total_enroll, 0.75), .(county_name)]
```

```{r}
# `model_data` contains a subset of the original merged data
# convert percentage to basis points for easy interpretation
model_data <- merged[year == 2016, .(county_name, school_name, school_cd, mean_ela_score, mean_math_score, county_per_poverty = county_per_poverty * 100)]
# keep only counties with enrollment below the 75th percentile
# after the merge, we have no use for `county_name`
model_data <- model_data[low_enroll_counties, on = "county_name"][, county_name := NULL]


# reshape the data from wide to long for plotting purposes
plot_data <- melt(model_data, id.vars = c("school_name", "school_cd", "county_per_poverty"))
scatter_plot(plot_data, aes(x = county_per_poverty, y = value, color = variable)) + 
  scale_color_manual(values = c("blue", "darkred"), labels = c("ELA", "math")) +
  labs(title = "County-level poverty v school test scores - 2016",
       x = "County-level poverty (%)",
       y = "Scores")
```
```{r}
model = lm(formula = mean_ela_score ~ county_per_poverty, data = model_data)
summary(model)
```


By removing schools from counties with large enrollment, our R2 has increased slightly from 0.04353 to 0.08812. This still means that **poverty rate only captures about 8% of variation in average test scores**. Because of this large variation in test scores across schools, it makes more sense to use a school-level variable like percentage of students with free or reduced lunch as a proxy for measuring poverty, as opposed to aggregating data for all schools across counties.

Bonus: does percentage of students with free and reduced lunch capture the variation in county poverty?

```{r}
# subset from `merged`, aggregate to country level
sum_table <- merged[ , .(tot_enroll = sum(total_enroll, na.rm = TRUE),
                         tot_reduced_lunch = sum(num_reduced_lunch, na.rm = TRUE),
                         tot_free_lunch = sum(num_free_lunch, na.rm = TRUE),
                         county_per_poverty = mean(county_per_poverty, na.rm = TRUE))
                     , county_name]

# convert totals to percentages
sum_table[, per_free_reduced_lunch := (tot_reduced_lunch + tot_free_lunch) * 100/ tot_enroll]

# remove the columns with total student numbers
sum_table[, tot_enroll := NULL][, tot_reduced_lunch := NULL][, tot_free_lunch := NULL]
head(sum_table)
```
```{r}
model = lm(formula = county_per_poverty ~ per_free_reduced_lunch, data = sum_table)
summary(model)
```
> 70% of the variation in county poverty level is captured by the average enrollment in free and reduced lunch programs across schools.
