---
title: "Final Exercise"
author: "Martin Pan"
date: "9/15/2020"
output: html_document
---

## Load Library and Data

```{r}
library(plm)
library(tidyverse)
```

```{r}
nys_school <- read_csv('../data/nys_schools.csv')
nys_acs <- read_csv('../data/nys_acs.csv')
```

## Explore Data

### nys_school
```{r}
head(nys_school)
```


```{r}
str(nys_school)
```

```{r}
summary(nys_school)
```

```{r}
sapply(nys_school, function(x) sum(is.na(x)))
```

### nys_acs
```{r}
head(nys_acs)
```
```{r}
str(nys_acs)
```

```{r}
summary(nys_acs)
```

```{r}
sapply(nys_acs, function(x) sum(is.na(x)))
```

## Variable Manipulation

### Deal With Missing Variable

```{r}
sapply(nys_acs, function(x) sum(x==-99))
```


```{r}
sapply(nys_school, function(x) sum(x==-99))
```

We are going to drop those rows with missing values
```{r}
nys_school_select <- filter(nys_school, county_name != -99, mean_math_score != -99, 
                            mean_ela_score !=-99, per_free_lunch != -99)
```

```{r}
sapply(nys_school_select, function(x) sum(x==-99))
```
We will ignore the issue of missing variables for distrct_name for now.

### Create Categorical Group
We will group 25 percentile and below for county_per poverty to low, 25 percentile to 75 percentile as medium, and the rest as high.

```{r}
threshold <- quantile(nys_acs$county_per_poverty, c(0.25, 0.75))

mean_poverty_rates <- nys_acs %>%
  group_by(county_name) %>%
  summarise(mean_pov = mean(county_per_poverty)) %>%
  mutate(poverty_group = case_when(
    mean_pov < threshold['25%'] ~ 'low',
    mean_pov > threshold['75%'] ~ 'high',
    (mean_pov > threshold['25%']) & (mean_pov < threshold['75%']) ~"medium"
  )) %>%
  select(!mean_pov)

nys_acs <- left_join(nys_acs, mean_poverty_rates, by="county_name")
```
```{r}
head(nys_acs)
```

### Scale Test Score
```{r}
nys_school_select <- nys_school_select %>%
  group_by(year) %>%
  mutate(scale_math_score = scale(mean_math_score), 
         scale_ela_score = scale(mean_ela_score))
```

```{r}
names(nys_school_select)
```

## Merge Datasets
We are going to take the mean for each county for school information.
```{r}
county_school <- nys_school_select %>%
  group_by(county_name, year) %>%
  summarise_if(is.numeric, mean)
```

```{r}
head(county_school)
```
Now, we will inner join county and acs dataset by county and year.
```{r}
merge_county <- inner_join(county_school, nys_acs, by = c('county_name', 'year'))
```

```{r}
head(merge_county)
```
## Create summary tables

### For Each County
```{r}
county_info <- merge_county %>% 
  group_by(county_name) %>%
  summarise(total_enrollment = sum(total_enroll),
            free_lunch_percent = mean(per_free_lunch),
            poverty_percent = mean(county_per_poverty),
            mean_reading_score = mean(mean_ela_score),
            mean_math_score = mean(mean_math_score))
```
```{r}
county_info
```
### Top 5 and Bottom 5
```{r}
county_info_sort <- arrange(county_info, poverty_percent)
head(county_info_sort)
```
```{r}
tail(county_info_sort)
```

## Data Visualization

### Lunch and Test
```{r}
nys_school_select %>%
  mutate(access_to_lunch = per_free_lunch + per_reduced_lunch,
         test_score = scale_ela_score + scale_math_score) %>%
  ggplot() + 
  geom_point(aes(x = access_to_lunch, y= test_score)) +
  labs(title="relationship between free/reduced price lunch and test performance at school level")
```
### Test Performance Across Countys
```{r}
ggplot(data=merge_county) +
  geom_point(aes(x=scale_math_score, y=scale_ela_score, color=poverty_group)) + 
  labs(title='Comparison of Test Score betwen Different Poverty Group')
```


## Answering Questions

We are going to run a regular ols and fixed effect model to explore the relationship.

```{r}
merge_county <- mutate(merge_county, total_scale_score = scale_ela_score + scale_math_score)
head(merge_county)
```

```{r}
mod <- lm(total_scale_score ~ county_per_poverty + per_free_lunch, data=merge_county)
summary(mod)
```
Let's preprocess the data to run the fixed effect model. 
```{r}
id_df <- as.data.frame(cbind(unique(merge_county$county_name), 1:length(unique(merge_county$county_name))))
names(id_df) <- c('county_name', 'id')
merge_county <- left_join(merge_county, id_df, by='county_name')
```

Now we can run the model.
```{r}
panel_merge_county <- pdata.frame(merge_county, index = c("id","year"))
mod_fixed <- plm(total_scale_score ~ county_per_poverty + per_free_lunch, data=panel_merge_county, model="within")
summary(mod_fixed)
```

We think that the poverty rate of each county does not have statistically significant effect on test score.