---
title: "Junpeng Jiang 2020 Bootcamp Submission"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(ggplot2)
```

# Task 1 
```{r}
school <- read.csv(here::here("data/nys_schools.csv"))
acs <- read.csv(here::here("data/nys_acs.csv"))
```

# Task 2
```{r}
str(school)
str(acs)
summary(school)
summary(acs)
```

- Check for data types
- There missing values in school dataset are labeled as -99, all categorical variables (plus school_cd) are complete.


# Task 3
### Dealing with Missing values

- Deal with -99, first count how many missing values there are.

```{r}
num_of_na <- rep(NA,6)
for (i in 7:12) {
  num_of_na[i-6] <- sum(school[,i] == -99)
}
num_of_na
dim(school)
```

- Remember that there are 35663 rows in the school dataset, so taking out 2200 rows should not affect the entire dataset considerably.

```{r}
new_school <- school
new_school[new_school == -99] <- NA
new_school <- na.omit(new_school)
dim(new_school)
```

- This way we take out all rows with missing values, and the remaining dataset new_school has 33437 observations.

### Create poverty level.

```{r}
county_avg_median <- acs %>% 
  group_by(county_name) %>% 
  summarise(n = mean(median_household_income))
county_avg_median <- as.data.frame(county_avg_median)
hist(county_avg_median$n)
```

- Compute average median income of each county across the years and plot a hisgram of the averages.

```{r}
summary(county_avg_median$n)
```

- By looking at the quantiles, I decide to split the groups as follows:
- < 46731 : low
- between 46731 and 56768: medium
- larger than 56768:high

```{r}
county_avg_median$group <- rep(NA,dim(county_avg_median)[1])
for (i in 1:dim(county_avg_median)[1]) {
  if(county_avg_median[i,'n'] < 46731){
      county_avg_median[i,'group'] <- "low"
  } else if(county_avg_median[i,'n'] >= 46731 && county_avg_median[i,'n'] <= 56768){
    county_avg_median[i,'group'] <- "medium"
  } else if(county_avg_median[i,'n'] > 56768){
    county_avg_median[i,'group'] <- "high"
  }
}
head(county_avg_median)
```

- This dataframe indexes each county with a group of low/medium/high.
- Then append the group variable back to the county dataset.

```{r}
acs$group <- rep(NA,dim(acs)[1])

for (i in 1:dim(acs)[1]) {
  rnum <- which(acs$county_name[i] == county_avg_median$county_name)
  acs[i,"group"] <- county_avg_median[rnum,"group"]
}
head(acs,15)
```

- The group variable is created as stated.

### Create Z-Score
```{r}
z_dt <- new_school %>% group_by(year) %>% summarise(n = scale(mean_ela_score), m = scale(mean_math_score))

new_school$z_ela <- z_dt$n
new_school$z_math <- z_dt$m
head(new_school)
```

# Task 4
```{r}
total_dt <- merge(school,acs,by = c("county_name","year"))
head(total_dt)
```

# Task 5

- 1

```{r}
tb1 <- total_dt %>% 
  group_by(county_name) %>% 
  summarise(total_erm = sum(total_enroll),
            q_for_lunch = sum(total_enroll*per_free_lunch)+sum(total_enroll*per_reduced_lunch),
            per_poverty = mean(county_per_poverty))

tb1 <- as.data.frame(tb1)
tb1$q_for_lunch <- tb1$q_for_lunch/tb1$total_erm
head(tb1)
```

- 2
```{r}

for_names <- tb1 %>% arrange(per_poverty)
c_names <- for_names$county_name[c(1:5,58:62)]
c_names# These are the counties has top5 and bottom5 poverty rate.

tb2 <- total_dt %>% 
  filter(county_name %in% c_names) %>% 
  group_by(county_name) %>% 
  summarise(q_for_lunch = sum(total_enroll*per_free_lunch)+sum(total_enroll*per_reduced_lunch),
            per_poverty = mean(county_per_poverty),
            avg_read = mean(mean_ela_score),
            avg_math = mean(mean_math_score),
            total_erm = sum(total_enroll))
tb2 <- as.data.frame(tb2)
tb2$q_for_lunch <- tb2$q_for_lunch/tb2$total_erm
tb2$total_erm <- NULL
head(tb2)
```

# Task 6
```{r}
length(unique(school$school_name))
plt1 <- new_school %>% 
  filter(year == 2008) %>% 
  group_by(school_name) %>% 
  summarise(access = per_free_lunch+per_reduced_lunch,
            total_mean_score = mean_ela_score+mean_math_score)
plt1 %>% ggplot() +
  geom_point(aes(x=access,y = total_mean_score))

plt2 <- new_school %>% 
  filter(year == 2009) %>% 
  group_by(school_name) %>% 
  summarise(access = per_free_lunch+per_reduced_lunch,
            total_mean_score = mean_ela_score+mean_math_score)
plt2 %>% ggplot() +
  geom_point(aes(x=access,y = total_mean_score))

plt3 <- new_school %>% 
  filter(year == 2010) %>% 
  group_by(school_name) %>% 
  summarise(access = per_free_lunch+per_reduced_lunch,
            total_mean_score = mean_ela_score+mean_math_score)
plt3 %>% ggplot() +
  geom_point(aes(x=access,y = total_mean_score))
```

- By subsetting data from 2008-2010, I plotted the scatter plot of the total percentage of access to free/reduced price lunch versus the total mean scores of each school.

- Through the plot we can see a negative linear trend: for schools from which students have more access to free/reduced price lunch, these students tend to have lower total scores on both ELA and math exams.

```{r}
plt22 <- total_dt %>% 
  group_by(group,year) %>%
  summarise(mean_score = (mean(mean_ela_score) + mean(mean_math_score))/2)

plt22 <- as.data.frame(plt22)
plt22 %>% ggplot +
  geom_line(aes(x = year,y = mean_score,group = group,col = group))
```

- This plot shows the average scores of each poverty level group across the years.
- If we take averages across all counties with their group labels (high/low/medium), the differences between each group's mean score are not as considerable.

# Task 7

```{r}
t7 <- total_dt %>% 
  group_by(county_name) %>% 
  summarise(mean_score = mean(mean_ela_score+mean_math_score),
            total_per = mean(per_free_lunch+per_reduced_lunch),
            group = group[1])
t7 <- as.data.frame(t7)

t7 %>% ggplot() +
  geom_point(aes(x = mean_score,y = total_per,group = group,col = group))

```

- In the plot, each point represents a county, and the colors represent their poverty labels. 

- From the plot we cannot see clear relationship between each counties poverty level(measured by the total percentage of students allowed for free/reduced price lunch) and their students' test performances.

- However, this plot is using data that is county-level, which means that the test scores and lunch data are averaged across all schools within one county. If we go back to the plot 1 in Task 6, we do see that the schools with lowers average scores often allow more students to access free/less expensive lunch.

- Also, this trend has been consistend across the years, as illustrated by the three plots in Task 6.


