---
title: "Day 5:  Advanced data manipulation 1 - dplyr"
author: "Richard Paquin Morel and Emilio Lehoucq"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
params:
  notes: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 4.35, fig.width = 4.75, message = FALSE, warning = FALSE)
```

```{r, include=FALSE}
notes<-params$notes
```

## dplyr - an idiosyncratic approach to data manipulation

- `dplyr` is an R package for data manipulation developed by Hadley Wickham with help from lots of other R developers
- https://dplyr.tidyverse.org/
- Principles: consistent syntax, meaningful naming conventions, fewer arguments/more functions 

## `dplyr` verbs

- `select` --> subset variables/columns
- `filter` --> subset observations/rows based on conditions
- `mutate` --> add new variables and keep existing ones
- `transmute` --> add new variables and drop existing ones
- `summarize` --> reduce multiple observations to a single value (e.g., find the mean)


## Trying out `dplyr` verbs - `select`

- select by name: 
  - `select(gapminder, continent, pop)`
- select by position: 
  - `select(df, c(1, 3, 10))`
- select by range:
  - `select(df, country:pop)` or `select(df, 1:3)`
- drop variables with `-`
  - `select(df, -gdpPercap)`

## Reload previous data 

```{r importing}
library(here)
library(readr)
library(reshape2)
library(lubridate)

generation <- read.csv(here::here("data/ca_energy_generation.csv"), stringsAsFactors=F)
imports <- read.csv(here::here("data/ca_energy_imports.csv"), stringsAsFactors=F)
generation$datetime <- as_datetime(generation$datetime)
imports$datetime <- as_datetime(imports$datetime)

merged_energy <- merge(generation, imports, by = "datetime")
long_merged_energy <- melt(merged_energy, id.vars = "datetime",
                           variable.name = "source",
                           value.name = "usage")

```

## Trying out `dplyr` verbs - `select`
```{r select}
library(dplyr)
tmp <- select(merged_energy, biogas, biomass, geothermal, solar)
names(tmp)
tmp <- select(merged_energy, -biogas, -biomass, -geothermal, -solar)
names(tmp)
```

## select helpers

- There are several "select helpers" that make subsetting variables very easy
  - `one_of()`, `contains()`, `starts_with()`, `ends_with()`, `matches()`

```{r}
tmp <- select(merged_energy, contains("hydro"), starts_with("bio"))
names(tmp)
```

## `filter`

```{r filter}
tmp <- filter(merged_energy, imports > 7000)
nrow(tmp)
head(tmp)
```

## multiple conditions in `filter`

```{r multi filter}
tmp <- filter(merged_energy, imports > 7000, natural_gas < 7000)
nrow(tmp)
head(tmp)
```

## `mutate`

- `mutate` creates new variables and keeps existing ones

```{r mutate}
tmp <- mutate(long_merged_energy, log_usage = log(usage))
head(tmp)
```

## `mutate`

```{r multi mutate}
tmp <- mutate(long_merged_energy, log_usage = log(usage), usage2 = usage^2, usage3 = usage^3)
head(tmp)
```

## `transmute`

- `transmute` creates new variables and keeps existing ones

```{r transmute}
tmp <- transmute(long_merged_energy, log_usage = log(usage))
head(tmp)
```

## `transmute`

```{r multi transmute}
tmp <- transmute(long_merged_energy, log_usage = log(usage), usage2 = usage^2, usage3 = usage^3)
head(tmp)
```

## `summarize`

`summarize` reduces observations to a single value based on functions
  - `mean`, `sum`, `sd`, `min`, `max`, etc.

```{r summarize}
# total energy consumption
summarize(long_merged_energy, total = sum(usage, na.rm = T))

# mean energy consumption
summarize(long_merged_energy, mean_cons = mean(usage, na.rm = T))
```

## A new operator: the pipe `%>%`

- The powerful `%>%` operator lets you chain together functions
- It sends the result of one function to another function
- Read `%>%` as "then"
- `%>%` increases the readability of your code
- Don't chain more than ~10 lines of code

```{r introducing the pipe}
# take df then filter it then select these variables
# you do not need to repeat the name of the dataframe!
long_merged_energy %>% 
  filter(source == "geothermal") %>% 
  select(-datetime) %>% 
  mutate(log_usage = log(usage)) %>% 
  summarize(mean_log_usage = mean(log_usage, na.rm = T))
```


## A new operator: the pipe `%>%`

- While piping, the piped dataframe _is not changed_!
- To refer to the manipulated dataframe, use `.`

```{r dot}
merged_energy %>% 
  select(-datetime) %>% 
  mutate(total_usage = rowSums(., na.rm = T)) %>% 
  summarize(total_usage = sum(total_usage, na.rm = T))
```

## Exercise 1

- Using the (wide) merged CA energy data `merged_energy` do the following with pipes:
  1. Select variables that contain the word "hydro"
  2. Create a new variable called `total_hydro` that is the sum of the retained hydro variables
  3. Find the mean usage for `total_hydro`

## Solution to exercise 1

```{r exercise}
merged_energy %>% 
  select(contains("hydro")) %>% 
  mutate(total_hydro = rowSums(., na.rm = T)) %>%
  summarize(mean_hydro = mean(total_hydro, na.rm = T))
```

## `group_by` and `summarize`

- `group_by` is a powerful function that allows us to perform operations by groups of observations!
  - Very useful with summarize!

```{r group_by}
long_merged_energy %>% 
  group_by(source) %>% 
  summarize(sum_usage = sum(usage, na.rm = T))
```

## `group_by` versus `for` loops

- Remember the `for` loop we wrote to find the mean life expectancy by year?
- MUCH better in `dplyr`

## `group_by` versus `for` loops

```{r mean le}
gapminder <- read.csv(here::here("data/gapminder5.csv"))

gapminder %>% 
  group_by(year) %>% 
  summarize(mean_le = mean(lifeExp, na.rm = T),
            sd_lf = sd(lifeExp, na.rm = T))
```

## Exercise 2

- Use your knowledge of `dplyr` to find the mean usage for small hydro, large hydro, biogas, and biomass
- Start with either the wide or long merged dataset

## Solution to exercise 2

```{r exercise grouping}
merged_energy %>% 
  select(datetime, contains("hydro"), contains("bio")) %>% 
  melt(id.vars = "datetime",
       variable.name = "source",
       value.name = "usage") %>% 
  group_by(source) %>% 
  summarize(mean_usage = mean(usage, na.rm = T))
```

## Merging in `dplyr` with `*_join` functions

- `dplyr`'s philosophy is to have more functions with fewer options
  - `left_join` - keeps all observations in the first dataframe 
    - `merge(all.x = T)`
  - `right_join` - keeps all observations in the second dataframe 
    - `merge(all.y = T)`
  - `full_join` - keeps all observations in both dataframes
    - `merge(all = T)`
  - `inner_join` - keeps only those observations that are matched in both datasets
    - `merge(all = F)`

## Comparing `merge` and `inner_join`
```{r merge and semi_join}
tmp <- merge(generation, imports, by = "datetime", all = F)
dim(tmp)
tmp <- inner_join(generation, imports, by = "datetime")
dim(tmp)
```

