---
title: "Exercises Part 6"
author: "Richard Paquin Morel, adapted from exercises by Christina Maimone"
date: "`r Sys.Date()`"
output: html_document
params:
    answers: TRUE
---


```{r, echo=FALSE, eval=TRUE}
answers<-params$answers
```

```{r global_options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(echo=answers, eval=answers,
                      warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE)
```

# Load the data

Load the `gapminder` dataset.

```{asis}
### Answer
```

```{r}
gapminder <- read.csv(here::here("data/gapminder5.csv"), stringsAsFactors=FALSE)
```


# If Statement

Use an if() statement to print a suitable message reporting whether there are any records from 2002 in the gapminder dataset. Now do the same for 2012.

Hint: use the `any` function.

```{asis}
### Answer
```

```{r}
year<-2002
if(any(gapminder$year == year)){
   print(paste("Record(s) for the year",year,"found."))
} else {
  print(paste("No records for year",year))
}
```


# Loop and If Statements

Write a script that finds the mean life expectancy by country for countries whose population is below the mean for the dataset

Write a script that loops through the `gapminder` data by continent and prints out whether the mean life expectancy is smaller than 50, between 50 and 70, or greater than 70.

```{asis}
### Answer
```

```{r}
overall_mean <- mean(gapminder$pop)

for (i in unique(gapminder$country)) {
   country_mean <- mean(gapminder$pop[gapminder$country==i])
   
   if (country_mean < overall_mean) {
     mean_le <- mean(gapminder$lifeExp[gapminder$country==i])
     print(paste("Mean Life Expectancy in", i, "is", mean_le))
   } 
} # end for loop
```

```{r}
lower_threshold <- 50
upper_threshold <- 70
 
for (i in unique(gapminder$continent)){
    tmp <- mean(gapminder$lifeExp[gapminder$continent==i])
    
    if (tmp < lower_threshold){
        print(paste("Average Life Expectancy in", i, "is less than", lower_threshold))
    }
    else if (tmp > lower_threshold & tmp < upper_threshold){
        print(paste("Average Life Expectancy in", i, "is between", lower_threshold, "and", upper_threshold))
    }
    else {
        print(paste("Average Life Expectancy in", i, "is greater than", upper_threshold))
    }
   
}
```


# Exercise: Write Functions

Create a function that given a data frame will print the name of each column and the class of data it contains. An example test data frame is supplied below.  Hint: Use `mode()` to get the class of the data in each column.  

```{r, eval=TRUE, echo=TRUE}
testdf <- data.frame(val1=1:5, val2=c("A","B","C","D","E"), 
                     stringsAsFactors = FALSE)
```


Create a function that given a vector and an integer will return how many times the integer appears in the vector.

Create a function that given a vector will print the mean and the standard deviation, it will optionally also print the median.

```{asis}
### Answer

Note: Some of these were taken or modified from https://www.r-bloggers.com/functions-exercises/
```

```{r}
data_frame_info <- function(df) {
  for (col in names(df)) {
    print(paste0(col, ": ", mode(df[,col])))
  }
}
data_frame_info(testdf)
```


```{r}
appearances <- function(x, val) {
  sum(x == val)
}
appearances(c(1,2,2,3,3,3,4), 4)
```

```{r}
vector_info <- function(x, include_median=FALSE) {
  print(paste("Mean:", mean(x)))
  print(paste("Standard Deviation:", sd(x)))
  if (include_median) {
    print(paste("Median:", median(x)))
  }
}
vector_info(c(1,2,2,3,3,3,4))
vector_info(c(1,2,2,3,3,3,4), TRUE)
```

