---
title: "Day 3 Exercises: Loops, Conditionals, and Functions"
author: "Sarah Moore"
date: "8/20/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## `here` 

Use the here command to load in the gapminder5.csv
```{r}
library(here)
here()
gapminder<-read.csv(here::here("data", "gapminder5.csv"))
``` 

Inspect the dataset and the variable classes to make sure that they have the proper attributes (e.g. change factors to characters). 

```{r}
head(gapminder)
str(gapminder)
gapminder$country<-as.character(gapminder$country)
gapminder$continent<-as.character(gapminder$continent)
```
Create a vector for each i observation in the dataset. 
```{r}
obs<-1:nrow(gapminder)
```

## For Loops

Using a `for` command, create a variable that measures the life expectancy of each country-year unit in months. 
Call this variable `lifeExp_months
```{r}
for(i in obs){
  gapminder[i, "lifeExp_months"]<- gapminder[i,"lifeExp"]*12
}
gapminder$lifeExp_months
```

Using another `for` command, create a new variable that is the log value of `lifeExp_months`. 
Call this variable `log_lifeExp_months`.
```{r}
for(i in obs){
  gapminder[i, "log_lifeExp_months"]<- log(gapminder[i,"lifeExp_months"])
}
gapminder$log_lifeExp_months
```
Use the vectorized method to replicate both of these variables without a loop. 
Call them lifeExp_months_2 and log_lifeExp_months_2, respectively. Compare these to the looped method. 
```{r}
gapminder$lifeExpmonths_2<-gapminder$lifeExp*12

gapminder$log_lifeExp_months_2<-log(gapminder$lifeExp*12)
#or 
gapminder$log_lifeExp_months_2<-log(gapminder$lifeExpmonths_2)

all(gapminder$lifeExp_months== gapminder$lifeExpmonths_2)
all(gapminder$log_lifeExp_months== gapminder$log_lifeExpmonths_2)
```
Using a method of your choosing, use the gdpPercap variable and pop variable to calculate the GDP per year per country. Call this variable "gdp". 
```{r}
for (i in obs) {
  gapminder[i, "gdp"]<- gapminder[i, "pop"]*gapminder[i, "gdpPercap"]
}

#or 

gapminder$gdp<-gapminder$pop*gapminder$gdpPercap
```

Now account for each of the unique country names that are in the dataset, and create a variable with them called "countries" 
```{r}
countries<-unique(gapminder$country)
``` 

Create a nested loop in which you create a readout of the country name and a calculation of the average log gdp for all years accounted for in the dataset. 

Repeat this for the standard deviation of country log gdp's over time. 
```{r}
for (i in countries){
  mean_gdp<-mean(log(gapminder$gdp)[gapminder$country==i], 
                 na.rm = T)
  print(paste0(i, ": ", mean_gdp))
}

for (i in countries){
  sd_gdp<- sd(log(gapminder$gdp)[gapminder$country==i], 
                  na.rm = T)
  print(paste0(i, ": ", sd_gdp))
}
```


## While Loops 

Set an interator starting with 1954, use a while loop to calculate the standard deviation of log gdp for ever 3 years before 2002 (inclusive). 

Make the read out so that it reads as "Year: standard deviation" (hint: see lecture notes for this particular format)
```{r}
gapminder$log_gdp<-log(gapminder$gdp)
  
i<-1954
while (i <= 2002) {
  sd_gdp_post1980<-sd(gapminder$log_gdp[gapminder$year==i], na.rm = T)
  print(paste0(i, ": ", sd_gdp_post1980))
  i<-i+3
}
```

## Conditional Statements 

Then create an `if/else` command that prints "Mean gdp per capita is less than 900" if the mean gdp per capita is less than 900 over all observed years. Else, print the statement "Mean GDP per capita is greater than 900." 

```{r}
if(mean(gapminder$gdpPercap) < 900) {
  print(paste("Mean GDP per capita is less than", 900))
} else {
  print(paste("Mean GDP per capita is greater than", 900))
}
```

Create a `for` loop with an if/else conditional: 
Write a `for` loop that calculates the mean gdp per capita for each country over the observed years in the dataset. If a given country's mean gdp per capita is greater than the median for all country-year observations, then print the country and its mean gdp per capita. `Else`, print the country name and the statement "Sorry, this country does not meet the median threshold." 

```{r} 
for(i in unique(gapminder$country)) {
  mean_gdpPercap<- mean(gapminder$gdpPercap[gapminder$country==i], na.rm=T)
  
  if(mean_gdpPercap > median(gapminder$gdpPercap, na.rm=T)){
  print(paste0(i, ": ", mean_gdpPercap))
  
} else { 
  print(paste0(i, ": Sorry, this country does not meet the median threshold."))
  }
}
```

## Functions 

Create a function called `report_min_max` to calculate the minimum and maximum log gdp (i.e. not per capita) for Europe. 

```{r}
report_min_max<-
  function(df, variable, continent){
    var<-gapminder[[variable]][gapminder$continent==continent]
    min_log_gdp<-min(log(var))
    max_log_gdp<-max(log(var))
    cat("Continent", continent, 
        "\nMinimum Log GDP", min_log_gdp, 
        "\nMaximum Log GDP", max_log_gdp)
  }

report_min_max(gapminder, "gdp", "Europe")
```

## Putting things all together. 

Change the following function to conform to the following:
#1) Change the color to green (hint: the specification for color is `col`)
#2) Instead of a log x variable, use a squared x variable. 
#3) Run the function as we did in the exercise from the lecture, but with the above new specifications for the year 1987. 
```{r}
viz_lm <-
  function(df, dv, iv, year) {
    dat <- df[df[["year"]] == year, ]
    y <- log(dat[[dv]])
    x <- (dat[[iv]])^2
    fit <- lm(y ~ x)
    plot(y ~ x, main = year,
         xlab = iv, ylab = dv)
    lines(x, predict(fit), col = 'green')
  }

viz_lm(gapminder, "lifeExp", "gdpPercap", 1987)
```

