---
title: "Data visualization with ggplot2	 - exercises"
author: "Emilio Lehoucq"
date: "`r Sys.Date()`"
output: html_document
params:
    answers: TRUE
---


```{r, echo=FALSE, eval=TRUE}
answers<-params$answers
```

```{r global_options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(echo=answers, eval=answers,
                      warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE)
```

```{r}
library(tidyverse)
```


# Load the data

Load the gapminder data.

```{asis}
### Answer 
```

```{r}
gapminder <- read.csv(here::here("data/gapminder5.csv"))
```

# Bringing skills together: histogram

1. Do a histogram of population

```{asis}
### Answer 
```

```{r}
ggplot(gapminder) +
    geom_histogram(aes(pop))
```

2. Adjust the number of bins to visualize the data more meaningfully.

```{asis}
### Answer 
```


```{r}
ggplot(gapminder) +
    geom_histogram(aes(pop), bins = 15)
```

3. Plot the proportion instead of the count on the y axis.

```{asis}
### Answer 
```

```{r}
ggplot(gapminder, aes(pop)) +
    geom_histogram(aes(y = (..count..)/sum(..count..)), bins = 20)
```

4. Change the titles for the axes, add a title for the graph, and add a caption for the source of the data

```{asis}
### Answer 
```

```{r}
ggplot(gapminder, aes(pop)) +
    geom_histogram(aes(y = (..count..)/sum(..count..)), bins = 20) +
      labs(title="Population of world countries", caption="Source: Gapminder", x="Population", y="Proportion of countries")
```

5. Center the title and write it in bold letter

```{asis}
### Answer 
```

```{r}
ggplot(gapminder, aes(pop)) +
    geom_histogram(aes(y = (..count..)/sum(..count..)), bins = 20) +
      labs(title="Population of world countries", caption="Source: Gapminder", x="Population", y="Proportion of countries") +
  theme(plot.title = element_text(hjust = 0.5, face="bold"))
```

6. Change the background to black and white and drop the border around the graph

```{asis}
### Answer 
```

```{r}
ggplot(gapminder, aes(pop)) +
    geom_histogram(aes(y = (..count..)/sum(..count..)), bins = 20) +
      labs(title="Population of world countries", caption="Source: Gapminder", x="Population", y="Proportion of countries") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5, face="bold"), panel.border = element_blank())
```

7. Display numbers in standard form rather than scientific notation

```{asis}
### Answer 
```

```{r}
library(scales)

ggplot(gapminder, aes(pop)) +
    geom_histogram(aes(y = (..count..)/sum(..count..)), bins = 25) +
      labs(title="Population of world countries", caption="Source: Gapminder", x="Population", y="Proportion of countries") +
    scale_x_continuous(labels = comma) +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5, face="bold"), panel.border = element_blank())
```

# Grouped data

1. Facet the histogram created in the previous histogram by continent.

```{asis}
### Answer 
```

```{r}
ggplot(gapminder, aes(pop)) +
    geom_histogram(aes(y = (..count..)/sum(..count..)), bins = 25) +
    labs(title="Population of world countries", caption="Source: Gapminder", x="Population", y="Proportion of countries") +
    facet_wrap(~continent) +
    scale_x_continuous(labels = comma) +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5, face="bold"), panel.border = element_blank())
```

2. Plot the evolution of the population per continent over time

```{asis}
### Answer 
```

```{r}
gapminder %>% 
    group_by(year, continent) %>% 
    mutate(population = sum(pop)) %>% 
    ggplot() +
    geom_line(aes(x = year, y = population)) +
    facet_wrap(~continent) +
    scale_y_continuous(labels = comma) +
    labs(title="Population of continents over time", caption="Source: Gapminder", x="Year", y="Population") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5, face="bold"), panel.border = element_blank())
```

```{r}
gapminder %>% 
    group_by(year, continent) %>% 
    mutate(population = sum(pop)) %>% 
    ggplot() +
    geom_line(aes(x = year, y = population, color = continent)) +
    scale_y_continuous(labels = comma) +
    labs(title="Population of continents over time", caption="Source: Gapminder", x="Year", y="Population") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5, face="bold"), panel.border = element_blank())
```

# Challenge

Reconstruct this graph.

```{asis}
### Answer 
```

```{r}
gapminder %>% 
    filter(country %in% c("China", "India", "Japan", "Hong Kong China", "Malaysia", "Taiwan", "Vietnam", "Indonesia", "Nepal")) %>% 
    ggplot(aes(gdpPercap, lifeExp)) +
    geom_point(size = 0.3) +
    geom_smooth(colour="black", size=0.5) +
    facet_wrap(~country) +
    labs(title="Life expectancy against GDP per capita in a sample of Asian countries", caption="Source: Gapminder", y="Life expectancy",
         x="GDP per capita") +
    theme_classic() +
    theme(plot.title = element_text(hjust = 0.5), panel.border = element_blank(), strip.background = element_rect(colour="white", fill="white"))
```
