---
title: "Data visualization with ggplot2	 - exercises"
author: "Emilio Lehoucq"
date: "`r Sys.Date()`"
output: html_document
params:
    answers: TRUE
---


```{r, echo=FALSE, eval=TRUE}
answers<-params$answers
```

```{r global_options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(echo=answers, eval=answers,
                      warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE)
```

# Load the data

Load the gapminder data.

```{asis}
### Answer 
```

```{r}
gapminder <- read.csv(here::here("data/gapminder5.csv"))
```

# Bringing skills together: histogram

1. Do a histogram of population

```{asis}
### Answer 
```

2. Adjust the number of bins to visualize the data more meaningfully.

```{asis}
### Answer 
```

3. Plot the proportion instead of the count on the y axis.

```{asis}
### Answer 
```

4. Change the titles for the axes, add a title for the graph, and add a caption for the source of the data

```{asis}
### Answer 
```

5. Center the title and write it in bold letter

```{asis}
### Answer 
```

6. Change the background to black and white and drop the border around the graph

```{asis}
### Answer 
```

7. Display numbers in standard form rather than scientific notation

```{asis}
### Answer 
```

# Grouped data

1. Facet the histogram created in the previous histogram by continent.

```{asis}
### Answer 
```

2. Plot the evolution of the population per continent over time

```{asis}
### Answer 
```

# Challenge

Reconstruct [this](https://github.com/MSIA/bootcamp-2020/blob/master/exercises/challenge_day7.png) graph.

```{asis}
### Answer 
```
