---
title: 'Day 1: Intro to R Syntax and Data Structures'
author: "Amanda Sahar d'Urso (materials from: K. Ramanathan)"
date: "`r Sys.Date()`"
output: 
  revealjs::revealjs_presentation:
    fig_height: 4
    self_contained: false
    reveal_plugins: ["notes"]
params:
  notes: no
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath("../"))
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
notes <- params$notes
```

# Introduction

## Why R?

> - R is flexible
> - R is customizable
> - R is open source
> - R has a large community of users

## RStudio and RStudio Server

> - RStudio refers to both the company and the software
> - RStudio is an integrated development environment (IDE) for R
> - RStudio Server is a browser-based version of RStudio that runs on a remote Linux server
> - We will be using RStudio today

<aside class="notes">
RStudio helps you organize your workflow and keep track of your work. 
The top-left pane is where you open, work on, and save script files. 
The bottom-left pane includes the console, which is where your code actually "runs" but is not saved.
The right-hand side panes include tools for manging your environment, workspace, and packages; for plotting and graphics; for accessing help files; and more.
</aside>

## Some key concepts

You will hear these terms:

> - Objects
> - Packages
> - Functions
> - Help files

# Basics

## First, best coding practices 

> - *always* use \`<-\` as your assignment mechanism, *not* \`=\`
> - *always* use spaces between operators
> - *never* have R save your environment upon exit (via global options)
> - *never* write code that, when shared, changes someone else's machine

## Don't do this: 

```{r not this example, eval = F}
install.packages("tidyverse")

y=1+1
```

## Do this:

```{r do this example, eval = F}
library(tidyverse)

y <- 1 + 1
```

## Arithmetic (R as a caluclator)

- `<-` is the assignment operator
- Standard arithmetic operators work: `+`, `-`, `/`, `*`, `^`
- To run an operation, you can hit Ctrl+Return (Windows) or Cmd+Return (Mac)

## Arithmetic (R as a calculator)

```{r arith 1}
y <- 5
y
y + 2
```

## Functions

- Functions take inputs and return outputs, e.g. `log(10)` results in `2.302585`
- Functions can also take arguments, e.g. `log(10, base=10)`
- All functions have a help file that you can call up, e.g. `?log`

## Functions 

```{r arith 2}
log(10)
log(10, base=10)
```

## Arithmetic: exercise

Open up the file `day1_syntax-fileIO_exercises.R` in RStudio. <br> Do the following tasks:

- Pick a number; save it as `x`
- Multiply x by 3
- Take the log of the above <br><small>(Hint, you need the function `log()` here)</small>
- Subtract 4 from the above
- Square the above


## Arithmetic: exercise

```{r arithmetic}
x <- 5
x*3
log(x*3)
log(x*3) - 4
(log(x*3) - 4)^2
```

## Comparisons and Logical Operators 

> - Common logical operators: 
    - `==` (is equal)
    - `!=` (not equal)
    - `<` (greater than), `>` (less than)
    - `&` (and), `|` (or)

> - If you evaluate statements with these operators, R will tell you if the statement is `TRUE` or `FALSE`

## Comparisons and Logical Operators: exercise

- Check if 1 is bigger than 2
- Check if 1 + 1 is equal to 2
- Check if it is true that the strings "eat" and "drink" are not equal to each other
- Check if it is true that 1 is equal to 1 *AND* 1 is equal to 2 <br><small>(Hint: remember what the operators `&` and `|` do)</small>
- Check if it is true that 1 is equal to 1 *OR* 1 is equal to 2

## Comparisons and Logical Operators: exercise

```{r comparisons}
1 > 2
(1 + 1) == 2
"eat" != 'drink'
(1 == 1) & (1 == 2)
(1 == 1) | (1 == 2)
```

## Packages

>- So far we have only used "base R". That is, the capabilities built into R.
>- What if you want to do other things in R? What about all the fancy things you've seen people do? Those require **packages** 
>- Load the package `tidyverse` using the function `library()` <br><small>You may need to install it using `install.packages("tidyverse")` if you have not done so already</small>
>- Open the help file for the function `recode` <br><small>(Hint: remember what `?` does)</small>

## Packages

```{r packages, eval=F}
library(tidyverse)

?recode
```

Remember, if you need to install a package:

```{r packages install, eval=F}
install.packages("tidyverse")
```

## Data types

> - **logical**: `TRUE` or `FALSE` <br><small>(When coerced into numeric type, `TRUE`=`1` and `FALSE`=`0`)</small>
> - **integer**: a specific type; most numbers are numeric instead
> - **numeric**: real or decimal 
> - **complex**: ex: `2+5i`
> - **character**: `"text data"`, denoted with quotes <br><small>(`'single quotes'` and `"double quotes"` both work)</small>
> - Use `typeof()` on an object to check its type

# Data structures

## Vectors

> - Vectors store multiple values of a single data type
> - Create a vector using the \`c()\` function
> - Vectors are *homogenous*: each vector can only contain one data type
> - Arithmetic operators and many functions can apply to vectors
> - Vectors can be *indexed* by:
>     - element position (`vec[1]`)
>     - 'slice' position (`vec[1:3]`)
>     - condition (`vec[vec>2]`).

## Vectors

```{r vectors example}
example <- c(7, 8, 9)
example[2]
example[example > 7]
```

## Vectors: exercise

Return to the exercise file and complete the following tasks:

- Run the code to generate variables `x1` and `x2`
- Select the 3rd element in `x1` <br><small>Hint: use `[]`</small>
- Select the elements of x1 that are less than 0
- Select the elements of x2 that are greater than 1
- Create x3 containing the first five elements of x2
- Select all but the third element of x1

## Vectors: exercise

```{r vectors}
set.seed(1234) 
x1 <- rnorm(5)
x2 <- rnorm(20, mean = 0.5)
x1[3]
x1[x1 < 0]
x2[x2 > 1]
x3 <- x2[1:5]
x1[-3]
```

## Interlude: Missing Values

Variables in datasets sometimes include missing variables. In R, missing values are stored as `NA`. Vectors containing any data type can contain missing values. Functions deal with missing values differently, and sometimes require arguments to specify how to deal with missing values.

```{r missing values}
vec <- c(1, 8, NA, 7, 3)
mean(vec)
mean(vec, na.rm = TRUE)
```

## Interlude: Missing Values

You can check if a vector contains missing values by the function `is.na()`. Since this returns a logical vector, you can use `sum()` or `mean()` on the result to count the number or proportion of `TRUE` values.

```{r missing values 2}
is.na(vec)
sum(is.na(vec))
mean(is.na(vec))
```

## Factors

> - Factors are a special type of vector that are useful for categorical variables
> - Factors have a limited number of *levels* that the variable can take, set by the user
> - For categorical variables with natural ordering between categories, we often want to use *ordered factors*
> - Create factors with `factor()`, which includes an argument for `levels = `

<aside class="notes">
We won't spend any more time reivewing factors today. If you need a refresher, check the lecture notes or review data camp. Each one tells you about factors from a different angle. The lecture notes focus on the utility of ordered factors, whereas data camp spends additional time on unordered factors.
</aside>

## Lists

> - Lists are like vectors, but more complex.
> - Lists are *heterogenous*: they can store single elements, vectors, or even lists.
> - You can keep multi-dimensional and ragged data in R using lists.
> - You can index an element in a list using double brackets: `[[1]]`. Single brackets will return the element as a list.
> - For more on lists, see the lecture notes or data camp.

<aside class="notes">
For more on lists, see the lecture notes or data camp.
</aside>

## Matricies

> - Matrices in R are two-dimensional arrays. 
> - Matrices are *homogenous*: all values of a matrix must be of the same data type. 
> - You can initialize a matrix using the `matrix()` function.
> - Matrices are used sparingly in R, primarly for numerical calculations or explicit matrix manipulation.
> - Matrices are indexed as follows: `mat[row no, col no]`.

<aside class="notes">
There is an example matrix for you to practice with in the exercises file, but we won't do it together.
</aside>

## Matricies

```{r matrix}
mat <- matrix(data = c(1,2,3,4,5,6,11,12,34), ncol = 3)
mat
mat[1,]
mat[,3]
mat[1, 3]
```

## Data frames

> - Data frames are the core data structure in R. A data frame is a list of named vectors with the same length.
> - Data frames are *heterogenous*: the vectors in a data frames can each be of a different data type.
> - Columns are typically variables and rows are observations.
> - You can make make data frames with `data.frame()`, or by combining vectors with `cbind()` or `rbind()`.
> - Data frames can be indexed in the same way as matricies: `df[row no, col no]`.
> - Data frames can also be indexed by using variable/column names: `df$var` or `df["var"]`.

## Data frames

```{r data frames example}
df <- data.frame(candidate = c("Biden","Warren","Sanders"), 
                 poll = c(26,17,17), 
                 age = c(77,71,78))
df
df[1,3]
df$age
```

## Data frames examples cont. 

```{r data frames examples cont}
df$candidate == "Biden"
df$age[df$candidate == "Biden"]
df$candidate[df$age == 76]
df$poll_max <- df$poll + 3
```

## Data frames: exercise

- Load the example data frame using the code provided
- Identify the number of observations (rows) and number of variables (columns)
- Identify the names of the variables
- Select the variable `mpg`
- Select the 4th row
- Square the value of the `cyl` variable and store this as a new variable `cylsq`

## Data frames: exercise

```{r data frames setup, include = F}
data(mtcars)
```

```{r data frames 1}
dim(mtcars) # str(mtcars) also okay here
names(mtcars)
```

```{r data frames 2, eval=F}
mtcars$mpg
```

```{r data frames 3}
mtcars[4,]
mtcars$cylsq <- (mtcars$cyl)^2
```

# The End!