---
title: "Advanced R - Reshaping, merging, and dplyr"
author: "Richard Paquin Morel"
date: "`r Sys.Date()`"
output: revealjs::revealjs_presentation
params:
  notes: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 4.35, fig.width = 4.75)
```

```{r, include=FALSE}
notes<-params$notes
```

# R Day 3 - Advanced manipulation and dplyr

## A new dataset

California energy data

- file names: `generation.csv` and `imports.csv`
- Read these two files into your working environment
  - They are in the "data" folder

## Reading in the data

```{r importing}
generation <- read.csv(here::here("data/generation.csv"), 
                       stringsAsFactors = F)
imports <- read.csv(here::here("data/imports.csv"), 
                    stringsAsFactors = F)
```

## Exploring the data

```{r explore data}
str(generation)
```

## Dealing with dates and times

- Notice that the first variable in both dataset is the called "datetime"
- What class are these variables?

```{r}
class(generation$datetime)
class(imports$datetime)
```

## Dealing with dates and times with `lubridate`

- The best way to deal with date-time data is to use the `lubridate` package and the `as_datetime` function
- Recode the `datetime` variable in the `imports` dataframe

```{r datetime}
library(lubridate)
generation$datetime <- as_datetime(generation$datetime)
class(generation$datetime)
```

## Dealing with dates and times with `lubridate`
```{r}
imports$datetime <- as_datetime(imports$datetime)
```

# Reshaping data

## Wide versus long

- Principles of "tidy data"

## Using `reshape2`

- `melt` --> make data long
- `dcast` --> make data wide
- `recast`--> melt then cast data

## Reshaping CA energy data

- Right now, the `generation` dataframe has several observations per row

```{r untidy data}
head(generation)
```

## `melt` the generation data

`melt(df, id.vars = "id")`
- Specify the variable that _doesn't_ melt with `id.vars`

```{r melting}
library(reshape2)
long_gen <- melt(generation, id.vars = "datetime")
head(long_gen)
```

# Merging data

## Merging CA energy data

- Sometimes you have data from two (or more) sources that you want to analyze 
- Need to merge these dataframes together
- To merge, need to chose the columns that have common values between the dataframes
  - Usually a variable with ids or years, or both
  
## Merging the `merge`

`merge(x, y, by = c("id", "year"))`
- Key arguments: 
  - `x`: first dataframe
  - `y`: second dataframe
  - `by`: variables to match (must have common name)

## More `merge` arguments
`merge(x, y, by.x = "id", by.y = "cd", all.x = T, all.y = T)`
- Advanced arguments:
  - Use `by.x` and `by.y` if the dataframes have different variable names
  - Use `all.x = T` if you want to keep all the observation in the first dataframe (unmatched observations in `y` are dropped!)
  - Use `all.y = T` if you want to keep all observations in the second dataframe (umatched observations in `x` are dropped!)
  - Use both (or, simply `all = T`) to keep all observations!
  - By **default** R will drop unmatched observations from **both** dataframe!

## Merge by id

```{r merge}
merged_energy <- merge(generation, imports, id = "datetime")
```

## Always be sure to check your merge!

```{r check merge}
dim(merged_energy)
head(merged_energy)
```

## Try reshaping the merged data!

- Our merged dataframe is still wide -- create a long version called `long_merged_energy`

## Try reshaping the merged data!

```{r melt exercise}
long_merged_energy <- melt(merged_energy, id.vars = "datetime",
                           variable.name = "source",
                           value.name = "output")
head(long_merged_energy)
```

# dplyr!

## dplyr - an idiosyncratic approach to data manipulation

- `dplyr` is an R package for data manipulation developed by Hadley Wickham with help from lots of other R developers
- https://dplyr.tidyverse.org/
- Principles: consistent syntax, meaningful naming conventions, fewer arguments/more functions 

## `dplyr` verbs

- `select` --> subset variables
- `filter` --> subset observations based on conditions
- `mutate` --> add new variables
- `summarize` --> reduce multiple variables to a single value (e.g., find the mean)


## Trying out `dplyr` verbs - `select`

- `select` subsets variables (columns)
  - select by name: `select(df, variable_1, variable_3, variable_10)`
  - select by position: `select(df, c(1, 3, 10))`
  - select by range: `select(df, variable_1:variable_10)` or `select(df, 1:10)`
  - drop variables with `-`

```{r select}
tmp <- select(merged_energy, biogas, biomass, geothermal, solar)
names(tmp)
tmp <- select(merged_energy, -biogas, -biomass, -geothermal, -solar)
names(tmp)
```

## select helpers

- There are several "select helpers" that make subsetting variables very easy
  - `one_of`, `contains`, `starts_with`, `ends_with`, `matches`

```{r}
tmp <- select(merged_energy, contains("hydro"), starts_with("bio"))
names(tmp)
```

## Trying out `dplyr` verbs - `filter`

```{r intro to verbs}
tmp <- filter(long_merged_energy, output > 1000)
nrow(tmp)
```

## Trying out `dplyr` verbs - `mutate`

- `mutate` creates new variables

```{r mutate}
tmp <- mutate(long_merged_energy, log_output = log(output))
head(tmp)
```

## Trying out `dplyr` verbs - `summarize`

`summarize` reduces observations to a single value based on functions
  - `mean`, `sum`, `sd`, `min`, `max`, etc.

```{r summarize}
# total energy consumption
summarize(long_merged_energy, total = sum(output, na.rm = T))

# mean energy consumption
summarize(long_merged_energy, mean_cons = mean(output, na.rm = T))
```

## A new operator: the pipe `%>%`

- The powerful `%>%` operator lets you chain together functions
- It sends the result of one function to another function
- Read `%>%` as "then"

```{r introducing the pipe}
# take df then filter it then select these variables
# you do not need to repeat the name of the dataframe!
long_merged_energy %>% filter(source == "geothermal") %>% select(-datetime) %>% mutate(log_output = log(output)) %>% summarize(mean_log_output = mean(log_output, na.rm = T))
```

## Try piping

- Using the (wide) merged CA energy data do the following with pipes:
  1. Select variables that contain the word "hydro" (and retain the datetime variable too!)
  2. Create a new variable called `total_hydro` that is the sum of the two hydro variables
  3. Find the mean output for hydro

## Try piping

```{r exercise}
merged_energy %>% 
  select(datetime, contains("hydro")) %>% 
  mutate(total_hydro = large_hydro + small_hydro) %>% 
  summarize(mean_hydro = mean(total_hydro, na.rm = T))
```

## `group_by` and `summarize`

- `group_by` is a powerful function that allows us to perform operations by groups of observations!
  - Very useful with summarize!

```{r group_by}
long_merged_energy %>% 
  group_by(source) %>% 
  summarize(sum_output = sum(output, na.rm = T))
```

## `group_by` versus `for` loops

- Remember the `for` loop we wrote yesterday to find the mean life expectancy by year?
- MUCH better in `dplyr`

## `group_by` versus `for` loops

```{r mean le}
gapminder <- read.csv(here::here("data/gapminder5.csv"))

gapminder %>% 
  group_by(year) %>% 
  summarize(mean_le = mean(lifeExp, na.rm = T),
            sd_lf = sd(lifeExp, na.rm = T))
```

## Find the mean by hour

- 

## Find the mean by hour

```{r exercise grouping}
merged_energy %>% 
  select(datetime, contains("hydro"), contains("bio")) %>% 
  mutate(total_hydro = large_hydro + small_hydro) %>% 
  melt(id.vars = "datetime",
       variable.name = "source",
       value.name = "output") %>% 
  group_by(source) %>% 
  summarize(mean_output = mean(output, na.rm = T))
```


## Merging in `dplyr` with `*_join` functions

- `dplyr`'s philosophy is to have more functions with fewer options
- `left_join` - keeps all observations in the first dataframe
- `right_join` - keeps all observations in the second dataframe 
- `full_join` - keeps all observations in both dataframes
- `inner_join` - keeps only those observations that are matched in both datasets

# Comparing `merge` and `inner_join`
```{r merge and semi_join}
tmp <- merge(generation, imports, by = "datetime")
dim(tmp)
tmp <- inner_join(generation, imports, by = "datetime")
dim(tmp)
```

