---
title: 'Intro to R: Day 3 - Data viz with ggplot'
author: "Kumar Ramanathan"
date: "`r Sys.Date()`"
output: 
  revealjs::revealjs_presentation:
    self_contained: false
    reveal_plugins: ["notes"]
params:
  notes: no
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath("../"))
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
notes<-params$notes
```

# Introduction to ggplot

Part of the `tidyverse`, `ggplot2` is an R package that implements the "grammar of graphics" in R.

In ggplot, graphics are built by supplying *data* and *mapping* of data values to aesthetics, and then adding *layers* that build geometries, scales, coordinate systems, and more.

## Why ggplot?

- Highly customizable and extensible.
- Designed to work with `dplyr` and other `tidyverse` packages.
- In use for over 10 years, meaning that help is widely available.
- Wide variety of avilable [extensions](http://www.ggplot2-exts.org/gallery/)

## ggplot vs. base R

# Basics: grammar of graphics

