---
title: 'Day 7: Data visualization with ggplot2'
author: |
  | Richard Paquin Morel
  | (slides created by Kumar Ramanathan and adapted by Emilio Lehoucq)
date: "`r Sys.Date()`"
output: 
  revealjs::revealjs_presentation:
    fig_height: 3.7
    self_contained: false
    reveal_plugins: ["notes"]
params:
  notes: no
height: 800
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath("../"))
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
```

```{r, include=FALSE}
notes<-params$notes
```

## ggplot2

Part of the `tidyverse`, `ggplot2` is an R package that implements the "grammar of graphics" in R.

Highly recommend videos of Hadley Wickham talking about it.

Definitely check out the lecture notes. They include many more things than the slides.

## Let's make sure we have the necessary data frames for our examples

Run the following code to have `imports`, `generation`, `merged_energy`, `long_gen`, `long_merged_energy`, `gapminder`, and `gapminder07` :

```{r, warning=FALSE, message=FALSE}
source(here::here("data/day3_objects.R"))
```

Or, if you want the option of using data.table syntax, also run this to convert the objects to data.tables. All the plotting will is the same, but you will need it for some data preparation.
```{r, warning=FALSE, message=FALSE}
library(data.table)
source(here::here("data/day3_objects_setDT.R"))
```

## Components of a basic plot

- **data**: a data frame, provided to the `ggplot()` function
- **geometric objects**: the objects/shapes that you want to plot, indicated through one of the many available `geom` functions, such as `geom_point()` or `geom_hist()`
- **aesthetic mapping**: the mapping from the data to the geometric objects, provided in an `aes()` function nested within the chosen `geom` function
- connected with the `+` operator

```{r ggplot structure, eval=F}
ggplot(data = <DATA FRAME>) + 
  <GEOM_FUNCTION>(mapping = aes(<VARIABLES>))
```

<aside class="notes">
Write this on the board, without data= and mapping= specified.
</aside>

## Example: Scatterplot from previous session

Task:<br>Create scatterplot of `lifeExp` against `gdpPercap` from `gapminder07` data.

Steps:

1. Supply `gapminder07` data frame as input to `ggplot()` function
2. Choose a `geom`; in this case we choose `geom_point()` to generate points
3. Supply aesthetic mapping, i.e. specify `x` and `y` values, to `geom_point()`.

## Example: Scatterplot from previous session

```{r scatterplot 1}
ggplot(data = gapminder07) + 
  geom_point(mapping = aes(x = gdpPercap, y = lifeExp))
```

## Example: Scatterplot from previous session

```{r scatterplot 2}
ggplot(gapminder07) + 
  geom_point(aes(x = gdpPercap, y = lifeExp)) + 
  labs(title = "Relationship between life expectancy and GDP per capita in 2007", 
       x = "GDP per capita", y = "Life expectancy")
```

## Exercise 1: Scatterplot

Using the `gapminder07` data, create a scatterplot of the natural log of `gdpPercap` as a function of the natural log of `pop`. Give it a title and axis labels.

<aside class="notes">
You can create a new script file for your exercises in this session, or you can continue to use script files from earlier in the day. In either case, make sure to give the file a sensible filename and to demarcate sections with meaningful comments.
</aside>

## Exercise 1: Scatterplot

```{r scatterplot exercise}
ggplot(gapminder07) + 
  geom_point(aes(x = log(pop), y = log(gdpPercap))) + 
  labs(title = "Relationship between GDP per capita and population in 
       2007", x = "Logged GDP per capita", y = "Logged life expectancy")
```

## Preparing your data

- Recall the role of pipes (`%>%`) in `dplyr` 
- `dplyr` and `ggplot2` are designed to work well together. 
- You can use a series of pipes to prepare your data before plotting.
- To see examples, let's turn to the California energy data.

## Example: Energy generated over time

Task:<br>Plot a *column chart* of *total* energy generated over time.

Steps: 

1. Choose long-format generation data, i.e. `long_gen`.
2. Manipulate data frame to calculate total output per date-time.
3. Pipe manipulated data frame into plot.
4. Select `geom_col()` and map appropriate `x` and `y` variables.

## Example: Energy generated over time

```{r data prep 2,message=FALSE}
# long_gen[,.(output = sum(output)), by = datetime] %>%
#   ggplot() + 
long_gen %>% 
  group_by(datetime) %>% 
  summarise(output=sum(output)) %>% 
  ggplot() + 
  geom_col(aes(x=datetime, y=output)) + 
  labs(title="Total energy generated, by hour", x="Hour", y="Output (MW)")
```

## Exercise 2: Hydro power generated over time

Task:<br>Plot a column chart *hydroelectric* power generated over time.

<small>Hint: There are two types of hydroelectric sources in the data: `large_hydro` and `small_hydro`.</small>

## Exercise 2: Hydro power generated over time

```{r data prep ex1,message=FALSE}
# long_gen[source %in% c("small_hydro", "large_hydro"),
#          .(output = sum(output)), by = datetime] %>%
#   ggplot() +
long_gen %>% 
  filter(source=="large_hydro"|source=="small_hydro") %>%
  group_by(datetime) %>% 
  summarise(output=sum(output)) %>%
  ggplot() + 
  geom_col(aes(x=datetime, y=output)) + 
  labs(title="Total hydro power generated, by hour", x="Hour", y="Output (MW)")
```

## Exercise 2: Hydro power generated over time

```{r data prep ex2}
# generation[,.(datetime, hydro = large_hydro+small_hydro)] 
#     ggplot() +
generation %>% 
  mutate(hydro=large_hydro+small_hydro) %>%
  ggplot() + 
  geom_col(aes(x=datetime, y=hydro)) + 
  labs(title="Total hydro power generated, by hour", x="Hour", y="Output (MW)")
```

## Geometric objects

We have already seen examples of two kinds of plots:

- Scatterplots, generated with `geom_point()`
- Column charts, generated with `geom_col()`

We'll see a few more, but there are *many* more! Your imagination is the limit!

## Line plots: `geom_line()`

```{r geom_line 1}
imports %>%
  ggplot() + 
  geom_line(aes(x=datetime, y=imports)) + 
  labs(title="Energy imports over time", x="Hour", y="Amount imported (MW)")
```

## Interlude: changing geom characteristics

- In the geom functions, we have only been supplying aesthetic mapping (in `aes()`).
- Geom functions can also control characteristics of the geom object, *outside* the `aes()` function.
- In most geoms, we can modulate `size=` and `col=`.
- Let's try it with the line plot we just made. 

## Interlude: changing geom characteristics

```{r geom_line 2}
imports %>%
  ggplot() + 
  geom_line(aes(x=datetime, y=imports), size=1.2, col="red") + 
  labs(title="Energy imports over time", x="Hour", y="Amount imported (MW)")
```

## Area plots: `geom_area()`

```{r area plot}
generation %>% 
  ggplot() + 
  geom_area(aes(x=datetime, y=wind), fill="darkblue") + 
  labs(title="Hourly wind power generation, Sept 3-9", x="Hour", y="Output (MW)")
```

## Box plots: `geom_boxplot()`

```{r box plot}
long_gen %>% 
  ggplot() + 
  geom_boxplot(aes(x=source, y=output)) + 
  labs(title="Amount of energy generated by each source, Sept 3-9", 
       x="Source type", y="Output (MW)")
```

## Multiple geoms in one plot

Task:<br>Plot a *line* of large hydro generation over time, and a *smoothed line* of the same relationship on top of it.

Steps: 

1. Supply the `generation` data to `ggplot()`.
2. Create a line with `geom_line()` with appropriate `x` and `y` aesthetics. Make it turquoise for fun.
3. Create a smoothed line with `geom_smooth()` with the same `x` and `y` aesthetics.

## Multiple geoms in one plot

```{r multiple geoms}
generation %>% 
  ggplot() + 
  geom_line(aes(x=datetime, y=large_hydro), col="turquoise3") + 
  geom_smooth(aes(x=datetime, y=large_hydro)) + 
  labs(title="Hydroelectric (large) generation per hour, Sept 3-9", 
       x="Hour", y="Output (MW)")
```

## Exercise 3: Total output per source

Task:<br>Create a column chart that shows the total output per source. 

- Change the color of the columns to `"darkred"`. 
- Add a horizontal line indicating the mean output across all sources. Use the cheatsheet to identify the `geom` function that you need.
- Add a meaningful title and axis labels using `labs()`.

## Exercise 3: Total output per source

```{r exercise3,message=FALSE}
# long_merged_energy[,.(output = sum(output)), by = source] %>%
#     ggplot() +
long_merged_energy %>%
  group_by(source) %>% 
  summarise(output=sum(output)) %>%
  ggplot() + 
  geom_col(aes(x=source, y=output), fill="darkred") + 
  geom_hline(aes(yintercept=mean(output))) + 
  labs(title="Total output per energy source over Sept 3-9", 
       y="Output (MW)", x="Source")
```

## Labels

```{r labels}
imports %>%
  ggplot() + 
  geom_line(aes(x=datetime, y=imports), col="red") + 
  labs(title="Energy imports over time in California", 
       subtitle="Hourly data from September 3-9, 2018",
       caption="Source: California Energy Commission", 
       x="Hour", y="Amount imported (MW)")
```

<aside class="notes">
No exercise this time, but we will practice this later.
</aside>

## Scales

- Scales allow you to manipulate the relationship between data values and aesthetics beyond aesthetic mapping.
- For example:
     - `scale_alpha`: control transparency settings
     - `scale_color` and `scale_fill`: control color palettes and other aspects of data mapped to colors
     - `scale_x` and `scale_y`: control the position of axis markers
- The cheatsheet is your friend!

## Scales

Task:<br>Recreate line plot of imports over time, but label x-axis with hours rather than dates.

Steps:

0. Check the cheatsheet!
1. Supply `imports` data to `ggplot()`.
2. Add a `geom_line()` layer and map appropriate `x` and `y` variables.
3. Add a `scale_x_datetime()` layer and use `date_labels=` and `date_breaks=` arguments to modify x-axis labels and breaks.
4. Add a `labs()` layer.

## Example: Imports over time

```{r scales dates}
imports %>%
  ggplot() + 
  geom_line(aes(x=datetime, y=imports), col="red") + 
  scale_x_datetime(date_labels="%H:%M", date_breaks="12 hours") +
  labs(title="Energy imports over time in California", 
       subtitle="Hourly data from September 3-9, 2018", 
       x="Hour", y="Amount imported (MW)")
```

<aside class="notes">
See lecture notes for details on what's going on inside the scale function. Or read the help file.
No exercise this time, but we will practice this later.
</aside>

## Themes: presets

ggplot comes with several preset themes that can be added as layers, including:

- `theme_grey()`: default
- `theme_bw()`: strip colors, including grey gradients
- `theme_dark()` and `theme_light()`: change the background of the coordinate system
- `theme_minimal()` and `theme_void()`: see reference guide for details

You can also manually alter many more things. Check out the help file.

## Example: Imports over time

```{r theme preset}
imports %>%
  ggplot() + geom_line(aes(x=datetime, y=imports), col="red") + 
  scale_x_datetime(date_labels="%H:%M", date_breaks="12 hours") +
  labs(title="Energy imports over time in California", 
       subtitle="Hourly data from September 3-9, 2018", 
       x="Hour", y="Amount imported (MW)") + 
  theme_minimal()
```

## Coordinate system adjustment

- `coord_flip()` is the default Cartesian coordinate system used; by explicitly calling this function, you can change the limits of the `x` and `y` axes from their defaults
- `coord_fixed()` sets a fixed aspect ratio between the `x` and `y` axes
- `coord_transform()` lets you transform the Cartesian coordinates using functions like `sqrt` or `log`
- `coord_polar()` changes the coordinate system to polar coordinates rather than a Cartesian system

## Example: coord_flip()

```{r coord_flip pre,message=FALSE}
# long_gen[,.(output = sum(output)), by = .(date = as.Date(datetime))] %>%
#     ggplot() +
long_gen %>% 
  mutate(date=lubridate::date(datetime)) %>% 
  group_by(date) %>% summarise(output=sum(output)) %>% 
  ggplot() + geom_col(aes(x=date, y=output)) + 
  labs(title="Total energy generated, by day", x="Day", y="Output (MW)")
```

## Example: coord_flip()

```{r coord_flip post,message=FALSE}
# long_gen[,.(output = sum(output)), by = .(date = as.Date(datetime))] %>%
#     ggplot() +
long_gen %>% 
  mutate(date=lubridate::date(datetime)) %>% 
  group_by(date) %>% summarise(output=sum(output)) %>% 
  ggplot() + geom_col(aes(x=date, y=output)) + 
  labs(title="Total energy generated, by day", x="Day", y="Output (MW)") + 
  coord_flip()
```

## Visualizing grouped data

- We often want to visualize multiple variables, or a variable that can be divided into groups, in one graphic.
- There are two broad principles for visualizing grouped variables in R:
     - Identify a grouping variable in a `geom()` using the `group = ` argument, and demarcate the distinct objects created per group in some way.
     - Use facets to generate geometric objects in separate coordinate systems for each group.

## Colors and fill

Task:<br>Create a line plot of energy output over time, with separate lines for each source.

Steps:

1. Supply the data frame `long_merged_energy` to `ggplot()`.
2. Add a `geom_line()` layer and specify `x=datetime` and `y=output`.
3. Also specify that `group=source` and `col=source`.
4. Add a `labs()` layer.

## Colors and fill

```{r aes colors 1}
long_merged_energy %>%
  ggplot() + 
  geom_line(aes(x=datetime, y=output, group=source, col=source)) + 
  labs(title="Output by energy source over time", 
       subtitle="Hourly data from September 3-9, 2018", x="Hour", 
       y="Output (MW)")
```

## Exercise 4: Colors and fill

Task:<br>Create a line plot that compares generation of wind, solar, and geothermal energy over time.

<small>Bonus: Set the line size to 1.5.</small>

## Exercise 4: Colors and fill

```{r aes colors ex}
# long_merged_energy[source %in% c("wind", "solar", "geothermal")] %>%
#     ggplot() +
long_merged_energy %>%
  filter(source=="wind"|source=="solar"|source=="geothermal") %>% 
  ggplot() + 
  geom_line(aes(x=datetime, y=output, group=source, col=source), size=1.5) + 
  labs(title="Wind vs. Solar vs. Geothermal generation", 
       subtitle="Hourly data from September 3-9, 2018", 
       x="Hour", y="Output (MW)")
```

## Customizing with scale_color layers

```{r aes colors 2}
# long_merged_energy[source %in% c("wind", "solar", "geothermal")] %>%
#     ggplot() +
long_merged_energy %>% filter(source=="wind"|source=="solar"| source=="geothermal") %>% 
  ggplot() + 
  geom_line(aes(x=datetime, y=output, group=source, col=source), size=1.5) + 
  scale_color_brewer(palette="Accent", name="Energy source") + 
  labs(title="Wind vs. Solar vs. Geothermal generation", 
       subtitle="Hourly data from September 3-9, 2018", 
       x="Hour", y="Output (MW)")
```

## Colors and fill

- `col=` is used to color objects like lines and points
- `fill=` is used to color objects like columns and histograms
    - `col=` will create colored outlines for such objects
- After some trial and error, this will become intuitive

## Position adjustment

 - Once we introduce a grouping variable to a `geom` function, we generate multiple geometric objects on the same plot.
 - Colors and fill help us distinguish each object from other objects.
 - Another way to distinguish objects, often used together with color/fill, is *position adjustment*.

## Example: Energy use by day

Task:<br>Create a column chart of energy use by day, grouped by source.

Steps:

1. Modify `long_merged_energy` to summarize output by date and source. Use `mutate()` to create a `date` variable from `datetime`. Use `group_by()` and `summarize()` to calculate total output per date per source.
2. Supply modified data frame to `ggplot()`
3. Add a `geom_col()` layer, supply `x` and `y` aesthetics.
4. Specify `group=source` and `fill=source`.
5. Add a `labs()` layer.

## Example: Energy use by day

```{r col stack,message=FALSE}
# long_merged_energy[,.(output = sum(output)), 
#                    by = .(source, date = as.Date(datetime))] %>%
long_merged_energy %>% 
  mutate(date=lubridate::date(datetime)) %>% 
  group_by(date, source) %>% 
  summarize(output=sum(output)) %>% 
  ggplot() + 
  geom_col(aes(x=date, y=output, group=source, fill=source)) + 
  labs(title="Energy use by day", x="Day", y="Output (MW)")
```

## Example: Energy use by day

- This column chart is useful if our main goal is to examine total output over time across sources.
- But what if our main goal is to compare trends across sources?
     - We can use `position="dodge"` in `geom_col()` 'unstack' the columns for each source.
- What if our goal is to compare what portion of output per day each source comprises?
     - We can use `position="fill"` to normalize height of stacked columns.

## Example: Energy use by day (pos dodge)

```{r col dodge,message=FALSE}
# long_merged_energy[,.(output = sum(output)), 
#                    by = .(source, date = as.Date(datetime))] %>%
long_merged_energy %>% 
  mutate(date=lubridate::date(datetime)) %>% 
  group_by(date, source) %>% 
  summarize(output=sum(output)) %>% 
  ggplot() + 
  geom_col(aes(x=date, y=output, group=source, fill=source), 
           position="dodge") + 
  labs(title="Energy use by day", x="Day", y="Output (MW)")
```

## Example: Energy use by day (pos fill)

```{r col fill,message=FALSE}
# long_merged_energy[,.(output = sum(output)), 
#                    by = .(source, date = as.Date(datetime))] %>%
long_merged_energy %>% 
  mutate(date=lubridate::date(datetime)) %>% 
  group_by(date, source) %>% 
  summarize(output=sum(output)) %>% 
  ggplot() + 
  geom_col(aes(x=date, y=output, group=source, fill=source), 
           position="fill") + 
  labs(title="Energy use by day", x="Day", y="Output (MW)")
```

## Shapes and linetypes

- Colors/fill and position adjustment are the most common and intuitive way to distinguish geometric objects from each other.
- Shapes and linetypes are two other ways.
- The idea here is to modify what the object actually looks like.

## Example: Energy source group, by day

Task:<br>

Steps:

1. Prepare data frame that summarizes output by date and "group" from `regroup`.
2. Supply modified data frame to `ggplot()`.
3. Two choices!
     1. Add `geom_line()` and `geom_point()` with `shape=group`.
     2. Add `geom_line()` with `linetype=group`.

## Example: Energy source group by day

```{r prep shapes,message=FALSE}
# Prepare data - data.table
# long_merged_energy[regroup, on = c(source = "type"), group := i.group]
# 
# long_merged_energy_regroup = 
#     long_merged_energy[,.(output = sum(output)), 
#                        by = .(group, date = as.Date(datetime))]

# Prepare data - dplyr
long_merged_energy_regroup <- long_merged_energy %>%
  rename(type = source) %>% 
  merge(regroup, by = "type") %>% 
  mutate(date=lubridate::date(datetime)) %>% 
  group_by(date, group) %>%
  summarise(output=sum(output))

head(long_merged_energy_regroup)
```

## Example: Energy source group, by day

```{r aes shapes}
long_merged_energy_regroup %>% 
  ggplot() + 
  geom_line(aes(x=date, y=output, group=group, col=group), size=0.8) +
  geom_point(aes(x=date, y=output, group=group, shape=group)) + 
  labs(title="Output by source group over time", 
       subtitle="Data collected during September 3-9, 2018", 
       x="Date", y="Output (MW)")
```

## Example: Energy source group, by day

```{r aes linetypes}
long_merged_energy_regroup %>% 
  ggplot() + 
  geom_line(aes(x=date, y=output, group=group, linetype=group), size=1) +
  labs(title="Output by source group over time", 
       subtitle="Data collected during September 3-9, 2018", 
       x="Date", y="Output (MW)")
```

## Sizes and alpha

- `size=` and `alpha=` inside `aes()` modulate the size and transparency of geom objects based on some data value.
- They are particularly useful when you want to distinguish objects based on some *continuous* variable.
- Let's return to the Gapminder data for an example.

## Example: Life expectancy over GDP

```{r aes size}
gapminder07 %>% 
  ggplot() + 
  geom_point(aes(x=log(gdpPercap), y=lifeExp, size=pop, col=continent)) + 
  scale_size_continuous(name="Population") + 
  scale_color_discrete(name="Continent") +
  labs(title="Life expectancy as a function of GDP per capita in 2007", 
       x="Logged GDP per capita", y="Life expectancy")
```

## Exercise 5: Average hourly output by source

Task:<br>Visualize the average output for each hour of the day, grouped by source.

<small>You need to identify the output per source per hour (e.g. 01:00, 02:00, etc) averaged over all days.</small>

- You will need to prepare your data using both `dplyr` and `lubridate` functions.
- You can choose which `geom`(s) to use, and how to demarcate groups.
- Bonus: use a scale layer to set a color palette (try `"Set3"`) and change the legend name.
- Remember to add `labs()`!

## Exercise 5: Average hourly output by source

```{r exercise5 setup,message=FALSE}
# ex5 <- long_merged_energy[,.(output = sum(output)), by = .(hour(datetime), source)] %>%
#     ggplot() +
ex5 <- long_merged_energy %>% 
  mutate(hour=lubridate::hour(datetime)) %>% 
  group_by(hour, source) %>% 
  summarize(output=sum(output)) %>%
  ggplot() + 
  geom_area(aes(x=hour, y=output, fill=factor(source))) + 
  scale_fill_brewer(palette="Set3", name="Source") + 
  labs(title="Average hourly output by source", 
       subtitle="Data collected during September 3-9",
       x="Hour of the day", y="Output (MW)") + 
  theme_bw()
```

## Exercise 5: Average hourly output by source

You can save images created by ggplot in three ways:

- Assign the result of your ggplot code to a named object, which can be saved like other objects in your R workspace
- Use the "save" icon in the Plots tab of RStudio
- Add the function `ggsave()` as a layer to your plotting code

## Exercise 5: Average hourly output by source

```{r exercise5 plot}
ex5
```

## Facets

- So far we have been visualizing grouped data by changing the appearance or position of the objects.
- A different approach is to use facets, which creates separate coordinate systems for each geometric object.

## Example: Comparing generation patterns

Task:<br>Compare energy generation over time, across sources.

How do we do this? Using what we've learned so far:

1. Supply `long_gen` to `ggplot()`.
2. Add a `geom_line()` layer, setting `x=datetime` and `y=output` in `aes()`.
3. Let's set `group=source` and `col=source` in `aes()`.

## Example: Comparing generation patterns

```{r why facet}
long_gen %>%
  ggplot() + 
  geom_line(aes(x=datetime, y=output, group=source, col=source), size=1) + 
  labs(title="Generation over time, by energy source", 
       subtitle="Hourly data from September 3-9, 2018", x="Hour", y="Output (MW)")
```

## Example: Comparing generation patterns

Is this a helpful plot?

- Not if our goal is to compare the patterns of each source.
- It's too noisy!
- Instead of setting `col=` in `aes()`, let's add a `facet_wrap()` layer.
- Specifically, `facet_wrap(~source)`, i.e. tell ggplot to "facet by source".

## Example: Comparing generation patterns

```{r facet 1}
long_gen %>% 
  ggplot() + 
  geom_line(aes(x = datetime, y = output)) + 
  facet_wrap(~source) + 
  labs(title="Generation over time, by energy source", 
       subtitle="Hourly data from September 3-9, 2018", 
       x="Hour", y="Output (MW)")
```

## Example: Comparing generation patterns

```{r facet 2}
long_gen %>% 
  ggplot() + 
  geom_line(aes(x = datetime, y = output)) + 
  facet_wrap(~source, scales="free") + 
  labs(title="Generation over time, by energy source", 
       subtitle="Hourly data from September 3-9, 2018", 
       x="Hour", y="Output (MW)")
```

## Exercise 6: Facets

Task:<br>Compare generation patterns for each source in facets. Color the lines using the "group" variable in `regroup`.

Remember:

- You will need to prepare your data. Think about the variables that you need: source/type, output, and group.
- When you pipe modified data into ggplot, remember that you are grouping in two ways: source/type through facets, and group through color.

## Exercise 6: Facets

```{r data prep for facets ex}
# data.table - don't need to create a new data.table
# long_gen[regroup, on = c(source = "type"), group := i.group]
# long_gen_regroup <- long_gen

# dplyr
long_gen_regroup <- long_gen %>%
  rename(type = source) %>% 
  merge(regroup, by="type")

head(long_gen_regroup)
```

## Exercise 6: Facets

```{r facets ex}
long_gen_regroup %>% ggplot() + 
  geom_line(aes(x=datetime, y=output, group=group, col=group), size=1) + 
  scale_color_brewer(palette="Set1", name="Type of energy source") +
  facet_wrap(~type, scales="free") + 
  labs(title="Generation over time, by energy source", 
       subtitle="Hourly data from September 3-9, 2018", 
       x="Hour", y="Output (MW)") + 
  theme(legend.position = "bottom")
```
